---
title: "LD-aware outlier regions: calling and benchmarking"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LD-aware outlier regions: calling and benchmarking}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 3.2, dpi = 100,
  cache = TRUE, autodep = TRUE       # cache chunks so re-knits skip the LD-decay / null steps
)
```

```{r packages}
library(LDscnR)
library(data.table)
library(ggplot2)
library(SNPRelate)
```

## The idea

Genome scans for selection produce noisy single-SNP signals. LDscnR turns them
into **outlier regions (ORs)** by leaning on two facts: true signal is
**LD-supported** (a causal variant drags a block of correlated markers with it),
and true outliers **cluster** in the genome. The pipeline scores each SNP by how
**consistently** it is both LD-prioritised and significant, calibrates a
threshold against a **structure-aware null**, and clusters the survivors into
regions.

The design collapses the usual thicket of tuning parameters to a principled few:

| role | parameter(s) | handled by |
|------|--------------|-----------|
| **swept / integrated** (natural [0,1]) | `rho` (`ld_w` window), `q*` (local-LD quantile) | folded into the C-score `ld_cscore()` |
| **fixed, principled** | `alpha` = 0.05 | the structure-aware null prices in any miscalibration |
| **clustering, LD-informed** | `rho_ld` = 0.9 (r²-link), 500 kb hard cap | `ld_edges()` |
| **operational knobs** | `tau_C` (null-calibrated), `l_min` (region size) | `calibrate_tauc()` / a post-filter |

The rest of the vignette shows both halves: **benchmarking** a scan against
simulated truth (`sim_ex`), and **calling** ORs on real data with a
null-calibrated threshold (`stickleback`).

## Part 1 -- benchmarking on simulated data

`sim_ex` is a small three-chromosome simulation with two QTNs and precomputed
EMMAX and LFMM F-statistics (`?sim_ex`). We turn the F-statistics into p-values,
flag the detectable true-positive QTNs, and build the LD ingredients.

```{r sim-setup, results = "hide"}
data(sim_ex, envir = environment())
GTs <- sim_ex$GTs
map <- copy(sim_ex$map)
colnames(GTs) <- map$marker   # align GTs colnames ("1:Pos") to map markers ("Chr1:Pos"), same order
n   <- nrow(GTs)

# F -> p (df 1, n-2); MAF and per-chromosome Va share for the truth flag
map[, emx_p  := pf(emx_F,  1, n - 2, lower.tail = FALSE)]
map[, lfmm_p := pf(lfmm_F, 1, n - 2, lower.tail = FALSE)]
af <- colMeans(GTs) / 2
map[, MAF := pmin(af, 1 - af)[marker]]
map[, Va  := p_Va]
map <- flag_true_qtns(map, va_col = "Va", maf_col = "MAF", maf_min = 0.1, p_va_min = 0.05)

# LD decay + a small ld_w window grid (the `rho` axis of the C-score)
gds <- create_gds_from_geno(GTs, map, tempfile(fileext = ".gds"))
set.seed(1)
ld_decay <- compute_LD_decay(gds, keep_el = TRUE, slide = 50, ld_method = "corr",
                             n_win_decay = 20, cores = 2)
rho_grid <- c(0.5, 0.7, 0.85, 0.95)
ld_ws <- sapply(rho_grid, function(r) compute_ld_w(ld_decay, rho = r, cores = 2))
rownames(ld_ws) <- unlist(lapply(ld_decay$by_chr, function(o) o$snp_ids), use.names = FALSE)
colnames(ld_ws) <- as.character(rho_grid)
ld_ws <- ld_ws[map$marker, ]
decay_sum <- ld_decay$decay_sum
```

### The C-score, clustering, and TP/FP

`ld_cscore()` integrates over the `ld_w` window (`rho`) and the local-LD quantile
(`q*`) at fixed `alpha`: a SNP's score is the fraction of cells where it is both a
local-LD candidate and FDR-significant. High-C SNPs are then clustered into ORs
(`ld_edges()` builds the r²-graph once; `ld_regions()` clusters any gated subset),
and scored against the QTN truth with `evaluate_ors()` (dedup-neutral: a duplicate
region for an already-claimed QTN is counted as neither TP nor FP).

```{r sim-score}
C_emx <- ld_cscore(map$emx_p, ld_ws)                       # per-SNP consistency
edges <- ld_edges(names(C_emx)[C_emx > 0], GTs, map, decay_sum, rho_ld = 0.9)
th    <- score_thresholds(decay_sum, rho_r2 = 0.75, rho_d = 0.95, dmax_cap = 5e5)
qtab  <- qtn_ld_table(GTs, map, names(C_emx)[C_emx > 0], max_bp = 2e6)

regions <- ld_regions(names(C_emx)[C_emx >= 0.3], edges)     # gate at tau_C = 0.3
regions <- regions[lengths(regions) >= 2]                    # l_min = 2
ev <- evaluate_ors(regions, map, qtab, th$r2min, th$dmax)
unlist(ev[c("TP", "FP", "Recall", "Precision")])
```

### C-score vs a single-SNP alpha scan

Because the C-score is one *ordered* knob (`tau_C`), a standard precision--recall
curve applies. We compare it against the single-SNP baseline (sweep `alpha` on the
raw p-values, same clustering and scoring), and summarise each with `pr_auc()`.

```{r sim-prauc}
pr_curve <- function(score_fun, grid) {
  rbindlist(lapply(grid, function(t) {
    reg <- ld_regions(score_fun(t), edges); reg <- reg[lengths(reg) >= 2]
    e <- evaluate_ors(reg, map, qtab, th$r2min, th$dmax)
    data.table(recall = e$Recall, precision = if (is.na(e$Precision)) NA_real_ else e$Precision)
  }))
}
cscore  <- pr_curve(function(t) names(C_emx)[C_emx >= t], seq(0.05, 1, 0.05))
singleS <- pr_curve(function(a) map$marker[p.adjust(map$emx_p, "BH") < a],
                    c(0.001, 0.005, 0.01, 0.05, 0.1))
c(C_score = pr_auc(cscore$recall, cscore$precision),
  single_SNP = pr_auc(singleS$recall, singleS$precision))
```

### Calibrating a second method against EMMAX (genomic control)

LFMM has no cheap structure-aware null. `gc_map_tauc()` transfers a threshold
calibrated on EMMAX onto LFMM's C-scale by quantile-matching -- genomic control on
the C-score, so LFMM is operated at EMMAX's FDR anchor.

```{r sim-gc}
C_lfmm  <- ld_cscore(map$lfmm_p, ld_ws)
tau_lfmm <- gc_map_tauc(tau_C = 0.3, C_ref = C_emx, C_target = C_lfmm)
signif(c(EMMAX_tau_C = 0.3, LFMM_tau_C = tau_lfmm), 3)
```

## Part 2 -- calling ORs on real data with a null-calibrated threshold

`stickleback` is a three-spined-stickleback subset with a marine/freshwater
ecotype phenotype and three regions (`?stickleback`): the **Chr1 inversion** and
the **Chr4 *Eda*** locus (positive controls), and **Chr6** background (negative
control). Here there is no ground truth, so instead of benchmarking we *call* ORs
with `ld_outlier_regions()`, which calibrates `tau_C` from the structure-aware
null.

The kinship should capture structure without absorbing the signal, so we build it
from **low-`ld_w` (LD-independent) markers** -- an all-marker GRM over-corrects and
erases localised signals like *Eda*.

We also raise the region-size filter to **`l_min = 10`**. `l_min` is a post-filter
(not part of the C-score) and it is the strongest, cleanest false-positive control
the pipeline has: a high-C **singleton** is almost always structure noise -- a lone
SNP that happens to be locally consistent -- whereas a true selected locus drags a
*block* of linked markers with it. And the structure-aware null essentially never
produces large clusters, so a &ge;10-SNP region is implausible by chance; raising
`l_min` therefore removes false regions far faster than true ones. The right value
scales with SNP density per LD block, so it is dataset-dependent: the sparse
`sim_ex` above used `l_min = 2` (its true regions span only a handful of SNPs),
while the dense stickleback data warrants `l_min = 10`.

```{r stickle-setup, results = "hide"}
data(stickleback, envir = environment())
sGTs <- stickleback$GTs
smap <- copy(stickleback$map)
eco  <- as.integer(stickleback$pheno$ecotype == "Marine")

sgds <- create_gds_from_geno(sGTs, smap, tempfile(fileext = ".gds"))
set.seed(1)
sdecay <- compute_LD_decay(sgds, keep_el = TRUE, slide = 200, ld_method = "corr",
                           n_win_decay = 50, cores = 2)
srho <- c(0.5, 0.7, 0.85, 0.95)
sld_ws <- sapply(srho, function(r) compute_ld_w(sdecay, rho = r, cores = 2))
rownames(sld_ws) <- unlist(lapply(sdecay$by_chr, function(o) o$snp_ids), use.names = FALSE)
colnames(sld_ws) <- as.character(srho); sld_ws <- sld_ws[smap$marker, ]

# neutral-background GRM: GCTA on the low-ld_w (LD-independent) markers
neutral <- smap$marker[sld_ws[, "0.95"] < 0.05]
K <- snpgdsGRM(sgds, snp.id = neutral, method = "GCTA", autosome.only = FALSE, verbose = FALSE)$grm
```

```{r stickle-call}
res <- ld_outlier_regions(
  y = eco, GTs = sGTs, K = K, ld_ws = sld_ws, map = smap, decay_sum = sdecay$decay_sum,
  basis = "genetic",              # confounding is genetic (ecotype ~ structure)
  rho_ld = 0.9, dcap = 5e5, l_min = 10, fdr = 0.05, B = 50, seed = 1
)
res
```

`ld_outlier_regions()` ran the whole pipeline -- fast EMMAX, C-score, a genetic
structure-aware null, `tau_C` calibrated to FDR ≤ 0.05, clustering and the `l_min`
filter -- and returned the surviving ORs with their chromosome, extent and peak C.
The positive-control regions (Chr1 inversion, Chr4 *Eda*) surface while Chr6
stays quiet.

Each surviving region gets its own colour from the package palette
`default_cluster_colours()`; SNPs not in a region are grey.

```{r stickle-plot, fig.height = 3}
reg_id <- rbindlist(lapply(seq_along(res$regions), function(i)
  data.table(marker = res$regions[[i]], region = as.character(i))))
d <- data.table(marker = names(res$C), C = as.numeric(res$C))[smap, on = "marker"]
d <- reg_id[d, on = "marker"][is.na(region), region := "ns"]

pal <- setNames(rep(default_cluster_colours(), length.out = length(res$regions)),
                as.character(seq_along(res$regions)))
ggplot() +
  geom_point(data = d[C > 0 & region == "ns"], aes(Pos, C), colour = "grey75", size = 0.6) +
  geom_point(data = d[region != "ns"], aes(Pos, C, colour = region), size = 1.2) +
  geom_hline(yintercept = res$tau_C, linetype = 2, colour = "grey40") +
  facet_wrap(~ Chr, nrow = 1, scales = "free_x") +
  scale_colour_manual(values = pal, guide = "none") +
  labs(x = "Position", y = "C-score",
       title = sprintf("%d outlier regions at tau_C = %.2f (l_min = 10)",
                       length(res$regions), res$tau_C)) +
  theme_bw(base_size = 10)
```

## Summary

- **`ld_cscore()`** integrates the two [0,1] nuisance parameters (`rho`, `q*`) at
  fixed `alpha` into one per-SNP consistency score.
- **`ld_edges()` / `ld_regions()`** cluster the survivors with an LD-decay-informed
  r² link and a 500 kb cap; **`l_min`** is a region-size post-filter.
- **`structured_null()` / `calibrate_tauc()`** set `tau_C` truth-free by pricing in
  structure confounding; **`gc_map_tauc()`** transfers it to a second method.
- **`ld_outlier_regions()`** chains all of the above into one call for real data.
- The **`evaluate_ors()` / `pr_auc()`** benchmark scores any caller against
  simulated truth on a common OR frame -- a method-agnostic gold standard.

See `?ld_outlier_regions` and the individual function help pages for the full
argument set.
