---
title: "LDscnR: LD-aware Genome Scans and Robust Outlier Detection"
output: rmarkdown::html_vignette
author: "Petri Kemppainen"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
message = FALSE,
warning = FALSE,
cache = TRUE,
comment = "#>",
fig.width = 7,
fig.height = 5
)

library(LDscnR)
#devtools::load_all()
data(LDscn)
data(ld_w_0.9)
data(ORs_tbl)
```

# 1. Introduction

Genome scans and genotype-environment association methods are widely used to detect loci under selection, but by ignoring correlations among loci (linkage disequilibrium, LD) they typically have either low power or high false positive rates under strong population structure and complex demographic histories. Both selective sweeps and reduced effective recombination and gene flow between locally adapted populations can lead to locally elevated levels of LD. Thus, contrasting local LD patterns against background levels caused by population structuring and complex demographic history can help focus inference on genomic regions more likely to be under selection. This perspective naturally motivates treating clusters of correlated SNPs (outlier regions, ORs) as test units rather than individual markers, reducing dimensionality while retaining biologically meaningful signal. Defining ORs relies on single linkage clustering (SLC) with threshold values for how strongly correlated physically adjacent outlier SNPs should be to be considered a single coherent outlier region. To account for differences in background LD and LD-decay rates, these thresholds are expressed relative to empirically estimated values.

LDscnR performs LD-aware scaling and quantile correction of genome scan statistics. The F-distribution provides a convenient and flexible reference family for one-degree-of-freedom tests commonly used in genome scans. The method is, however, agnostic to the specific test statistic used, and only assumes:

1.  **Monotonicity**. Larger values of the statistic $S_i$ correspond to stronger evidence against the null hypothesis.

2.  **Known null distribution**. Under the null hypothesis, $S_i$ can be mapped to a theoretical distribution with known quantiles.

3.  **Independence of null distribution from LD correction**. The theoretical null distribution is defined independently of the LD-weighting procedure.

Thus, the LD-scaling algorithm does not depend on the interpretation of $S_i$ itself, but rather on its null quantile structure.

Because both LD-scaling and OR definition depend on multiple tuning parameters, and because different methods capture complementary aspects of genomic differentiation, this framework integrates evidence across parameter values and inference methods.

This method is based on Kemppainen & Guillaume (2026a), but with speed and memory optimised functions and the additional improvements outlined in Kemppainen & Guillaume (2016b).

LDscnR provides a framework to:

1.  Estimate LD-structure,
2.  Estimate LD-decay parameters,
3.  Perform LD-weighted scaling of test statistics,
4.  Correct quantiles via permutation,
5.  Evaluate robustness across LD window scales and clustering parameters,
6.  Detect consistent outlier regions (ORs).

------------------------------------------------------------------------

## Transformation to the F-scale

LDscnR expects statistics expressed as:

$$
F_i \sim F(df_1, df_2)
$$

under the null hypothesis. Many commonly used genome scan statistics satisfy this condition either directly or via transformation:

| Original statistic   | Transformation to F-scale             |
|----------------------|---------------------------------------|
| p-value              | $F = F^{-1}(1 - p)$                   |
| $\chi^2_1$ statistic | $\chi^2_1 = F(1, \infty)$             |
| $t$-statistic        | $t^2 \sim F(1, df)$                   |
| $z$-score            | $z^2 \sim \chi^2_1 \sim F(1, \infty)$ |
| Likelihood ratio     | asymptotically $\chi^2 \rightarrow F$ |

Thus, for most genome scan statistics, the transformation route is:

$$
\text{statistic} \rightarrow p\text{-value} \rightarrow F\text{-statistic}.
$$

Once transformed, the statistic can be passed to `LDscn_pipeline()`, a wrapper function that performs the entire pipeline from estimating LD-structure (`compute_ld_structure()`) and LD-decay (`ld_decay()`), performing quantile transformation and detecting outlier regions (`ld_rho_draws()`) and generating consistency scores as final output (`consistency_score()`).

## Quick start: running an LD-scaled genome scan

We load an included simulated example that contain the necessary data for this pipeline.

```{r}
library(LDscnR)
data(sim_ex)
```

The dataset contains:

-   `GTs`: genotype matrix ($n_{\text{ind}} \times n_{\text{SNP}}$)
-   `map`: SNP metadata and test statistics that includes at least one F-based (pre-transformed, if necessary) and optionally `fdr` corrected p-values for the original test (q-values).
-   `env`: The environmental variable to be used in other optional outlier methods (not required for `LDscn_pipeline()`).

For completeness, in the below example include q-values as well.

```{r eval=FALSE}
LDscn <- LDscn_pipeline(geno = sim_ex$GTs,
                        map = sim_ex$map,
                        F_cols=c("emx_F","lfmm_F"),
                        q_cols=c("emx_q","lfmm_q"),
                        cores=8)
```

The object `LDscn` is created that contains the following slots:

1.  **Decay**. Contains information about LD-decay and background LD
2.  **ld_struct**. Edge lists of pairwise LD-values ($r^2$) per chromosome.
3.  **draws**. Outlier regions F-scaled statistics based on a range of parameter values.
4.  **consistency**. Consistency based statistics.
5.  **call**. Parameter settings used.

Next we can plot and inspect LD-structure.

```{r}
# Plots LD-decay and fitted line for one chromosome at a time
plot_ld_decay(LDscn,chr = "Chr1")

# Zoomed in
plot_ld_decay(LDscn,max_dist = 5e5) 

# Plot only d vs r^2 but for one chromosome at a time
plot(LDscn$ld_struct,chr = "Chr1")

# Plot fitted LD-decay curves for all chromosomes
plot(LDscn$decay)

```
# rests (not ready yeat)
## Next we want to add the $F´$-statistics and C-scores to the map file

\# 3. LD-decay estimation

LDscnR operates on GDS files via SNPRelate.

```{r, eval=FALSE}
gds_path <- tempfile(fileext = ".gds")
gds <- create_gds_from_geno(geno = GTs, map = map, gds_path)

decay  <- ld_decay(gds)
ld_str <- compute_ld_structure(gds, slide_win_ld = 500)
```

The LD-decay model is:

$$
  r^2(d) = b + \frac{c - b}{1 + a d},
$$

where:

-   $b$ is background LD,
-   $a$ controls decay rate,
-   $d$ is physical distance.

------------------------------------------------------------------------

\# 4. LD-weighted scaling

For a chosen LD-decay quantile $\rho_w$, LDscnR computes an LD-weighted statistic:

$$
  F' = F \cdot \ell_w,
  $$ where $\ell_w$ is a local LD weight derived from the decay model.

Quantile correction is applied using permutation of LD weights, aligning the empirical LD-inflated null to the theoretical $F$-distribution.

```{r, eval=FALSE}
scan <- ld_scan(
  ld_str,
  decay,
  SNP_ids = map$marker,
  F_vals  = map[, .(emx_F, lfmm_F)],
  rho_w   = 0.95,
  n_inds  = n_inds,
  full    = TRUE
)
```

Visualize quantile decomposition:

```{r, eval=FALSE}
plot(scan, method = "lfmm_F")
```

The QQ decomposition separates:

-   Theoretical null,
-   LD-induced distortion,
-   Raw LD-weighted statistic,
-   LD-scaled statistic.

------------------------------------------------------------------------

# 5. Robustness across LD window scales

Instead of fixing $\rho_w$, we evaluate robustness across a range:

```{r, eval=FALSE}
draws <- ld_rho_draws(
  ld_struct = ld_str,
  decay_obj = decay,
  F_vals    = map[, .(emx_F, lfmm_F)],
  q_vals    = map[, .(emx_q, lfmm_q)],
  n_inds    = n_inds,
  SNP_ids   = map$marker,
  n_rho     = 3,
  rho_w_lim = list(min = 0.8, max = 0.95),
  n_or_draws = 5,
  rho_d_lim  = list(min = 0.8, max = 0.95),
  rho_ld_lim = list(min = 0.9, max = 0.99),
  alpha_lim  = list(min = 1.5, max = 2),
  lmin_lim   = list(min = 1, max = 5)
)
```

------------------------------------------------------------------------

# 6. Consistency score

The consistency score (C-score) measures:

$$
C_i = \frac{\text{Number of draws where SNP } i \text{ is in an OR}}{\text{Total draws}}.
$$

```{r, eval=FALSE}
C_obj <- consistency_score(draws, combine = TRUE)
map   <- add_consistency_to_map(map, C_obj)
```

Interpretation:

-   $C_i \approx 1$ indicates highly robust signal.
-   $C_i \approx 0$ indicates instability across LD scales.

------------------------------------------------------------------------

# 7. Outlier region detection

Outlier regions are defined as connected components under LD constraints:

-   Distance threshold: $$
    d < d_\rho
    $$
-   LD threshold: $$
    r^2 > b + (1 - b)(1 - \rho_{\text{ld}})
    $$

Detection based on C-score:

```{r, eval=FALSE}
ORs <- detect_or(
  q_vals   = map[, .(C_mean)],
  ld_struct = ld_str,
  decay_obj = decay,
  sign_th   = 0.5,
  sign_if   = "greater",
  rho_d     = 0.9,
  rho_ld    = 0.95,
  l_min     = 2
)
```

Convert to table and merge:

```{r, eval=FALSE}
OR_tbl <- or_to_table(ORs)
map    <- OR_tbl[map, on = c("SNP" = "marker")]
map[is.na(OR_id), OR_id := "ns"]
```

------------------------------------------------------------------------

# 8. Visualization

Prepare Manhattan layout:

```{r, eval=FALSE}
layout <- prep_manhattan(
  map[, .(
    bp = Pos,
    Chr,
    marker,
    C_log = -log10(1 - C_mean + 1e-8),
    OR_id
  )],
  spacer = 0
)
```

Plot C-score landscape:

```{r, eval=FALSE}
plot_manhattan_gg(
  layout,
  y = "C_log",
  sig_th = -log10(1 - 0.5)
)
```

------------------------------------------------------------------------

# 9. Interpretation

LDscnR decomposes genome scan signals into:

1.  LD-induced distortion,
2.  True statistical signal,
3.  Robust spatial signal.

LD-scaling reduces structured noise from autocorrelation. The C-score quantifies robustness across LD window definitions. Outlier regions identify coherent clusters under explicit LD constraints.

------------------------------------------------------------------------

# 10. Reproducibility

```{r, eval=FALSE}
sessionInfo()
```

------------------------------------------------------------------------

# Summary

LDscnR provides an LD-aware framework for:

-   Statistical correction,
-   Robustness evaluation,
-   Spatial clustering,
-   Interpretable visualization.

It allows genome scans to be interpreted under explicit modeling of linkage disequilibrium.
