---
  title: "LDscnR: LD-aware Genome Scans and Robust Outlier Detection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LDscnR: LD-aware Genome Scans}
%\VignetteEngine{knitr::rmarkdown}
%\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>",
fig.width = 6,
fig.height = 5
)
```

# 1. Introduction

Genome scans based on $F$-statistics frequently exhibit **LD-induced inflation**, where correlated markers inflate test statistics and generate extended peaks of apparent signal.

LDscnR provides a framework to:

1.  Estimate LD-decay parameters,
2.  Perform LD-weighted scaling of test statistics,
3.  Correct quantiles via permutation,
4.  Evaluate robustness across LD window scales,
5.  Detect consistent outlier regions (ORs).

This vignette demonstrates the complete workflow on a simulated dataset.

------------------------------------------------------------------------

\# 2. Example dataset

We load an included simulated example:

```{r}
library(LDscnR)
library(SNPRelate)
library(data.table)

data(sim_ex)
```

The dataset contains:

-   `GTs`: genotype matrix ($n_{\text{ind}} \times n_{\text{SNP}}$)
-   `SNP_res`: SNP metadata and test statistics

We prepare the map:

```{r}
map <- copy(sim_ex$SNP_res[, .(
Chr, Pos, marker, type,
emx_F, lfmm_F
)])

GTs    <- sim_ex$GTs
n_inds <- nrow(GTs)
```

Compute FDR-adjusted q-values:

```{r}
map[, emx_q  := p.adjust(pf(emx_F,  1, n_inds - 2, lower.tail = FALSE), "fdr")]
map[, lfmm_q := p.adjust(pf(lfmm_F, 1, n_inds - 2, lower.tail = FALSE), "fdr")]
```

------------------------------------------------------------------------

\# 3. LD-decay estimation

LDscnR operates on GDS files via SNPRelate.

```{r}
gds_path <- tempfile(fileext = ".gds")
gds <- create_gds_from_geno(geno = GTs, map = map, gds_path)

decay  <- ld_decay(gds)
ld_str <- compute_ld_structure(gds, slide_win_ld = 500)
```

The LD-decay model is:

$$
  r^2(d) = b + \frac{c - b}{1 + a d},
$$

where:

-   $b$ is background LD,
-   $a$ controls decay rate,
-   $d$ is physical distance.

------------------------------------------------------------------------

\# 4. LD-weighted scaling

For a chosen LD-decay quantile $\rho_w$, LDscnR computes an LD-weighted statistic:

$$
  F' = F \cdot \ell_w,
  $$ where $\ell_w$ is a local LD weight derived from the decay model.

Quantile correction is applied using permutation of LD weights, aligning the empirical LD-inflated null to the theoretical $F$-distribution.

```{r}
scan <- ld_scan(
  ld_str,
  decay,
  SNP_ids = map$marker,
  F_vals  = map[, .(emx_F, lfmm_F)],
  rho_w   = 0.95,
  n_inds  = n_inds,
  full    = TRUE
)
```

Visualize quantile decomposition:

```{r}
plot(scan, method = "lfmm_F")
```

The QQ decomposition separates:

-   Theoretical null,
-   LD-induced distortion,
-   Raw LD-weighted statistic,
-   LD-scaled statistic.

------------------------------------------------------------------------

# 5. Robustness across LD window scales

Instead of fixing $\rho_w$, we evaluate robustness across a range:

```{r}
draws <- ld_rho_draws(
  ld_struct = ld_str,
  decay_obj = decay,
  F_vals    = map[, .(emx_F, lfmm_F)],
  q_vals    = map[, .(emx_q, lfmm_q)],
  n_inds    = n_inds,
  SNP_ids   = map$marker,
  n_rho     = 3,
  rho_w_lim = list(min = 0.8, max = 0.95),
  n_or_draws = 5,
  rho_d_lim  = list(min = 0.8, max = 0.95),
  rho_ld_lim = list(min = 0.9, max = 0.99),
  alpha_lim  = list(min = 1.5, max = 2),
  lmin_lim   = list(min = 1, max = 5)
)
```

------------------------------------------------------------------------

# 6. Consistency score

The consistency score (C-score) measures:

$$
C_i = \frac{\text{Number of draws where SNP } i \text{ is in an OR}}{\text{Total draws}}.
$$

```{r}
C_obj <- consistency_score(draws, combine = TRUE)
map   <- add_consistency_to_map(map, C_obj)
```

Interpretation:

-   $C_i \approx 1$ indicates highly robust signal.
-   $C_i \approx 0$ indicates instability across LD scales.

------------------------------------------------------------------------

# 7. Outlier region detection

Outlier regions are defined as connected components under LD constraints:

-   Distance threshold: $$
    d < d_\rho
    $$
-   LD threshold: $$
    r^2 > b + (1 - b)(1 - \rho_{\text{ld}})
    $$

Detection based on C-score:

```{r}
ORs <- detect_or(
  q_vals   = map[, .(C_mean)],
  ld_struct = ld_str,
  decay_obj = decay,
  sign_th   = 0.5,
  sign_if   = "greater",
  rho_d     = 0.9,
  rho_ld    = 0.95,
  l_min     = 2
)
```

Convert to table and merge:

```{r}
OR_tbl <- or_to_table(ORs)
map    <- OR_tbl[map, on = c("SNP" = "marker")]
map[is.na(OR_id), OR_id := "ns"]
```

------------------------------------------------------------------------

# 8. Visualization

Prepare Manhattan layout:

```{r}
layout <- prep_manhattan(
  map[, .(
    bp = Pos,
    Chr,
    marker,
    C_log = -log10(1 - C_mean + 1e-8),
    OR_id
  )],
  spacer = 0
)
```

Plot C-score landscape:

```{r}
plot_manhattan_gg(
  layout,
  y = "C_log",
  sig_th = -log10(1 - 0.5)
)
```

------------------------------------------------------------------------

# 9. Interpretation

LDscnR decomposes genome scan signals into:

1.  LD-induced distortion,
2.  True statistical signal,
3.  Robust spatial signal.

LD-scaling reduces structured noise from autocorrelation. The C-score quantifies robustness across LD window definitions. Outlier regions identify coherent clusters under explicit LD constraints.

------------------------------------------------------------------------

# 10. Reproducibility

```{r}
sessionInfo()
```

------------------------------------------------------------------------

# Summary

LDscnR provides an LD-aware framework for:

-   Statistical correction,
-   Robustness evaluation,
-   Spatial clustering,
-   Interpretable visualization.

It allows genome scans to be interpreted under explicit modeling of linkage disequilibrium.
