---
title: "A quick introduction to LDscnR"
output:
  pdf_document:
    toc: true
    number_sections: true
latex_engine: xelatex
vignette: >
  %\VignetteIndexEntry{A quick introduction to LDscnR}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  cache = TRUE
)

```

```{r packages}
library(LDscnR)
library(data.table)
library(SNPRelate)
```

## Overview

`LDscnR` provides tools for estimating chromosome-specific linkage disequilibrium (LD) decay from genotype data. It fits per-chromosome decay curves, relates decay rate to chromosome size, and derives recommended LD window sizes for user-defined relative LD thresholds ($\rho$).

This vignette gives a compact end-to-end example using the bundled simulated dataset `sim_ex`. The workflow covers:

-   creation of a GDS object from genotype data,
-   estimation of LD decay,
-   visualization of decay summaries and window recommendations,
-   converting between relative LD thresholds, physical distance, and expected $r^2$.

## Example data

The example dataset contains:

-   `GTs`: a genotype matrix,
-   `map`: a SNP annotation table including genomic positions and simulation-specific metadata.

```{r data}
data("sim_ex")

map <- sim_ex$map
```

To use the genotype data throughout the workflow, we first create a temporary GDS file.

```{r gds}
gds_path <- tempfile(fileext = ".gds")
gds <- create_gds_from_geno(geno = sim_ex$GTs, map, gds_path)
```

## Estimating LD decay

The first step is to estimate chromosome-wise LD decay. This is done in two stages.

We first run a preliminary analysis on a subset of markers to obtain a rough estimate of the sliding window size needed for downstream LD calculations.

```{r ld-decay-pre-load,include=FALSE,cache=FALSE}
cores <- 8
dir.create("vignette_cache", showWarnings = FALSE)
ld_file_pre <- "vignette_cache/ld_decay_pre.rds"

if (file.exists(ld_file_pre)) {
  ld_decay_pre <- readRDS(ld_file_pre)
} else {

ld_decay_pre <- compute_LD_decay(
  gds,
  n_win_decay = 5,
  max_SNPs_decay = 5000,
  keep_el = FALSE,
  slide = 1000,
  cores = cores
)
  saveRDS(ld_decay_pre, ld_file_pre)
}
```

```{r ld-decay-pre, eval=FALSE}


ld_decay_pre <- compute_LD_decay(
  gds,
  n_win_decay = 5,
  max_SNPs_decay = 5000,
  slide = 1000,
  cores = cores
)
```

This gives a summary of LD-decay based on the subset of the data.

```{r ld-decay-pre-show}
ld_decay_pre
```

In this simulated example, chromosomes are relatively short, so only five decay windows per chromosome are used. For empirical data sets, the default settings are usually more appropriate.

The preliminary LD-decay fit can be used to select a more suitable sliding window for the full analysis. We can visualize LD-decay, and use the recommended new slide window for estimating LD.

```{r ld-decay-plot,fig.width=8, fig.height=6}
plot(ld_decay_pre, type = "summary")
plot(ld_decay_pre, type = "recommendation")
plot(ld_decay_pre, type = "recommendation", rho = 0.99)
plot(ld_decay_pre, type = "chr", chr = "Chr2")
```

Here we use the mean recommended SNP window corresponding to `rho = 0.99`.

```{r ld-decay-full-load, include=FALSE, cache=FALSE}

new_slide_window <- mean(ld_decay_pre$decay_sum$slide_snp_rho_0.99)

dir.create("vignette_cache", showWarnings = FALSE)
ld_file <- "vignette_cache/ld_decay.rds"

if (file.exists(ld_file)) {
  ld_decay <- readRDS(ld_file)
} else {

ld_decay <- compute_LD_decay(
  gds,
  n_win_decay = 5,
  max_SNPs_decay = Inf,
  keep_el = TRUE,
  slide = new_slide_window,
  cores = cores
)
  saveRDS(ld_decay, ld_file)
}

```

```{r ld-decay-full, eval=FALSE}

new_slide_window <- mean(ld_decay_pre$decay_sum$slide_snp_rho_0.99)

ld_decay <- compute_LD_decay(
  gds,
  n_win_decay = 5,
  max_SNPs_decay = Inf,
  keep_el = TRUE,
  slide = new_slide_window,
  cores = cores
)

```

The second run uses all SNPs and stores LD edge lists (`keep_el = TRUE`). Note that the edge list of pairwise $r^2$ values can be written to files (using `fwrite`/`fread`) instead, to save RAM, by specifying e.g. `el_data_folder = "./el/"`.

## Inspecting the decay parameters

The chromosome-wise decay parameters are stored in `decay_sum`. Each chromosome has an estimated decay rate `a`, background LD `b`, and short-range LD `c`.

```{r decay-sum}
ld_decay$decay_sum
```

## Relative LD thresholds, distance, and expected LD

Decay is summarised on a relative scale via the threshold $\rho$. Two helper functions convert between $\rho$, physical distance, and expected $r^2$:

-   `d_from_rho(a, rho)` returns the physical distance (bp) at which a given relative threshold $\rho$ is reached, for a chromosome with decay rate `a`.
-   `ld_from_rho(b, c, rho)` returns the expected $r^2$ corresponding to $\rho$.

For example, using the parameters of the first chromosome:

```{r rho-helpers}
chr1 <- ld_decay$decay_sum[1]

# physical distance (bp) at which rho = 0.99 is reached
d_from_rho(a = chr1$a, rho = 0.99)

# expected r^2 at rho = 0.99
ld_from_rho(b = chr1$b, c = chr1$c, rho = 0.99)
```

These conversions underpin the recommended window sizes reported in `ld_decay$recommendation`, which are expressed in SNP units for a range of target $\rho$ values.

## Summary

This vignette illustrated the core `LDscnR` workflow using simulated data. Starting from genotype and SNP metadata, we:

-   created a GDS object,
-   estimated chromosome-wise LD decay in two stages,
-   visualized decay summaries, per-chromosome fits, and window recommendations,
-   converted between relative LD thresholds, physical distance, and expected $r^2$.

For empirical data, the same general workflow applies.

```{r cleanup, include=FALSE}
SNPRelate::snpgdsClose(gds)
unlink(gds_path)
```
