---
title: "LD-aware outlier regions from your own scan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LD-aware outlier regions from your own scan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## What this does

You have run a genome scan and have a p-value per marker. You want the
**regions** that stand out, not a list of individual markers, and you want a
defensible significance statement for each one.

`ld_scan()` takes two things: the p-values from your observed data, and the
p-values from `B` permuted datasets scanned exactly the same way. It returns
LD-supported regions, each with a p-value obtained by asking how often a
surrogate produced something as strong *at that same locus*.

Nothing about your phenotype, covariates or association model enters the
pipeline — only p-values. EMMAX, LFMM, BayPass, a plain GLM: all work, and two
engines can be compared on identical footing.

## Where the C-score comes from

The consistency C-score is **not original to this package**. It was introduced by
Fang, Kemppainen, Momigliano & Merilä (2021), *Population structure limits
parallel evolution in sticklebacks*, Mol Biol Evol 38(10):4205–4221, which
defines it by name along with `tau_C = 0.05`, `l_min = 10`, a 500 kb
single-linkage cap, and the region-as-cluster-of-significant-loci definition.
**Cite that paper when you use the C-score.**

What differs here is the shape of the integration, not the idea: continuous over
`rho` and `q*` rather than over a menu of 144 analyses, and per-SNP rather than
tested at cluster level — which is what lets `tau_C` and `l_min` be swept as a
second, nested scale by `ld_region_c2()`.

## What you need

| Input | What it is |
|---|---|
| `p_obs` | one p-value per marker, from your scan |
| `p_perm` | the same, for each of `B` permuted datasets |
| `ld_ws` | local-LD weights, from `compute_ld_w()` |
| `map` | `marker`, `Chr`, `Pos` |
| `GTs` | genotype dosages, individuals x markers |
| `decay_sum` | per-chromosome LD decay, from `compute_LD_decay()` |

## The one decision that matters: how you permute

The surrogates *define* what "no signal" means, and that choice is yours, not
the software's.

Suppose samples come from four localities, each containing both ecotypes.
Shuffling the ecotype label **within** each locality holds the between-locality
composition fixed, so a surrogate can rebuild any signal that merely tracks
geography — and such a signal will fail the test. Shuffling **globally** breaks
that link, so the same clinal signal now looks real. Neither is wrong; they
answer different questions. Decide which one you mean, then say so — the
`basis` label is carried through every printed output for exactly this reason.

```{r}
# one surrogate: permute, re-run YOUR scan, keep the p-values
one_surrogate <- function() {
  eco_perm <- ave(ecotype, locality, FUN = sample)   # within-locality shuffle
  my_scan(eco_perm)                                  # same engine, same settings
}
p_perm <- replicate(200, one_surrogate(), simplify = FALSE)
```

`B = 200` is a reasonable default: the smallest p-value any region can attain is
`1/(1+B)`, so `B = 200` gives a floor of 0.005 and `B = 96` gives 0.0103. If you
compare two engines, give them the **same** `B`, or their p-values are not on the
same scale.

## Running it

```{r}
fit <- ld_scan(p_obs, p_perm, ld_ws, map, GTs, decay_sum,
               tau = 0.05, l_min = 3, rho_ld = 0.6,
               basis = "within-locality permutation", engine = "EMMAX")
fit
```

Memory note: `p_perm` may also be a `function(b)` that reads surrogate `b` from
disk. Each is reduced to a sparse C-score and discarded, so 200 whole-genome
surrogates cost a few MB rather than gigabytes.

## Read the gate first

```{r}
fit$gate
```

The gate reports, per surrogate, the **median** number of markers with `C > 0`
and of regions of at least `l_min` markers, against the observed counts. If the
surrogates approach the observed numbers, the null reproduces the very signal it
is supposed to be a null for, and **its p-values mean nothing**. `ld_scan()`
warns and `print()` says so, but the judgement is yours.

Read the `max_regions` and `frac_surr_any_region` columns next to the median.
Permutation nulls are often heavy-tailed — silent in four surrogates out of five
and explosive in the fifth — and a median alone reports that as "silent".

Report a null that fails the gate rather than dropping it. A disclosed
mis-specified null is a result; a quietly deleted one is a hole in the methods.

## The regions

```{r}
fit$regions$regions
```

Each region carries `s_R` (its summed C-mass), `p_R`, and `q_R` after
Benjamini–Hochberg across the discovered regions:

$$s_R = \sum_{j \in R} C_j, \qquad
  p_R = \frac{1 + \#\{b : s^{null}_{b,R} \ge s_R\}}{1 + B}$$

The test is **location-matched**, and that is what makes it work. A signal
confined to part of your sampled range need not be reproduced by a surrogate *at
its own locus*, so a genuinely range-restricted peak — which a genome-wide count
would charge to population structure — can still separate. It is also robust to a
globally noisy null, because excess null peaks are scattered and rarely land on
any one observed locus.

`q_R` is the multiple-testing correction this method reports. A Bonferroni
correction over every marker would double-count a burden already carried at the
C-score and region-size stages.

## When everything sits at the p-floor

With a permutation null, the regions worth reporting usually all land at
`1/(1+B)`. That is not a failure — `q_R` is telling you "not a structure
artifact" — but it cannot rank them, and it is worth saying so plainly rather
than implying a resolution the null does not have.

Ranking is a separate job:

```{r}
fit$c2$regions      # regions ordered by C-squared
```

`ld_region_c2()` re-runs the region test across a `(tau, l_min)` grid and scores
each locus by the fraction of *usable* cells — those producing any significant
region — in which it comes out significant. Normalising by the usable grid is
what defeats cherry-picking: a locus significant only in the single cell that
maximises the region count scores near zero, while one significant across the
grid scores near 1.

`C^(2)` is a **ranking, not a second significance test**. The grid cells are not
independent and no error rate is controlled by it.

## Step by step

`ld_scan()` is a convenience wrapper. The stages are separately callable, which
is what you want when surrogates are expensive and you intend to reuse them:

```{r}
null  <- ld_null_from_p(p_obs, p_perm, ld_ws, basis = "...", engine = "...")
edges <- ld_edges(null$universe, GTs, map, decay_sum, rho_ld = 0.6, dcap = 5e5)

ld_gate(null, edges, tau = 0.05, l_min = 3)
ld_region_scan(null, edges, tau = 0.05, l_min = 3)
ld_region_c2(null, edges)
```

To gate several nulls side by side, pass a **named list** and one shared edge
graph. Edge membership is a property of a marker pair, not of the marker set, so
a graph built over the union of their universes gives identical clustering and
makes the rows comparable:

```{r}
univ  <- unique(unlist(lapply(nulls, `[[`, "universe")))
edges <- ld_edges(univ, GTs, map, decay_sum, rho_ld = 0.6, dcap = 5e5)
ld_gate(nulls, edges, tau = 0.05, l_min = 3)
```

## Choosing tau and l_min

Don't agonise. `tau = 0.05` and `l_min = 3` are sensible defaults, and
`ld_region_c2()` exists precisely so the choice does not have to carry weight —
it integrates both away. Raising `l_min` is the strongest simple lever on
false positives: chance clusters rarely grow large.

## Relationship to `ld_outlier_regions()`

`ld_outlier_regions()` is the older, self-contained route: it runs its own EMMAX
scan, generates its own MVN surrogates, and calibrates a single `tau_C` against a
pooled count-FDR. It is still there and still works. Prefer `ld_scan()` when you
want to choose the engine, choose the surrogates, or compare two engines.

`null_fdr()`, `calibrate_tauc()` and `calibrate_lmin()` remain exported as
**diagnostics** for describing a null's clustering propensity. They are no longer
the recommended way to pick an operating point.
