---
title: "A worked example: LD decay and complexity reduction on stickleback data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A worked example: LD decay and complexity reduction on stickleback data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 6, dpi = 100
)
```

```{r packages}
library(LDscnR)
library(data.table)
library(ggplot2)
library(patchwork)
library(SNPRelate)
```

## Overview

The [quick introduction](LDscnR_quick_introduction.html) walks through the API on
the small simulated `sim_ex` data. This worked example applies the same LD-decay
and LD-complexity-reduction workflow to **real data** -- the bundled
`stickleback` set (`?stickleback`): three-spined stickleback
(*Gasterosteus aculeatus*) marine--freshwater ecotype genotype dosages from
[Fang et al. (2021)](https://doi.org/10.1093/molbev/msab144). It is deliberately
built from three chromosomal regions with contrasting linkage-disequilibrium
regimes:

- **Chr1** -- a chromosomal inversion (structural, strong long-range LD),
- **Chr4** -- the *Eda* marine--freshwater adaptation locus (a high-LD outlier),
- **Chr6** -- a background region with rapidly-decaying, low LD.

That contrast is what makes the LD-decay and complexity-reduction steps visible,
so it is a good data set to *see* what the methods do. Genotypes are dosages in
$[0, 2]$ (no missing data), 117 individuals.

```{r data}
data(stickleback)
GTs <- stickleback$GTs
map <- copy(stickleback$map)

region_lab <- c(Chr1 = "Chr1 inversion (high LD)",
                Chr4 = "Chr4 Eda locus",
                Chr6 = "Chr6 background (low LD)")
```

```{r gds}
gds_path <- tempfile(fileext = ".gds")
gds <- create_gds_from_geno(GTs, map, gds_path)
```

## LD decay and per-marker local LD support

`compute_LD_decay()` fits the per-chromosome decay curve
$r^2(d) = b + (c-b)/(1+ad)$, where the decay rate $a$ is estimated in a series of
overlapping windows along each chromosome (`n_win_decay` sets the nominal window
count; 50% overlap roughly doubles it). `compute_ld_w()` then measures, at
**marker resolution**, the local LD support `ld_w`: the median $r^2$ between a
marker and its neighbours within the $\rho = 0.95$ window.

```{r decay, results = "hide"}
set.seed(1)  # compute_LD_decay() subsamples SNP pairs for the background-LD fit
ld_decay <- compute_LD_decay(gds, keep_el = TRUE, slide = 200,
                             ld_method = "corr", n_win_decay = 50, cores = 2)

ld_w <- compute_ld_w(ld_decay, rho = 0.95, cores = 2)
ids  <- unlist(lapply(ld_decay$by_chr, function(o) o$snp_ids), use.names = FALSE)
map[, ld_w_095 := ld_w[match(marker, ids)]]
```

The tracks below overlay, per region, the per-marker `ld_w` (grey points) with the
windowed decay rate `a` (orange line), each min--max scaled to 0--1 so they share
an axis. Because `a` is a *rate*, **low `a` means slow decay = high LD**, so the
two are inversely related: `a` bottoms out exactly under the high-`ld_w` blocks.

```{r tracks-fn}
sc01 <- function(v) { v <- as.numeric(v); (v - min(v, na.rm = TRUE)) / diff(range(v, na.rm = TRUE)) }

track_panel <- function(decay_obj, ch) {
  lw <- map[Chr == ch & is.finite(ld_w_095)]
  lw <- data.table(x = lw$Pos / 1e6, y = sc01(lw$ld_w_095))
  d  <- as.data.table(decay_obj$by_chr[[ch]]$decay)[regime == "structured" & is.finite(a)]
  dw <- data.table(x = ((d$start + d$end) / 2) / 1e6, y = sc01(d$a))[order(x)]
  ggplot() +
    geom_point(data = lw, aes(x, y), colour = "grey55", size = 0.25, alpha = 0.35) +
    geom_line(data = dw, aes(x, y), colour = "#D55E00", linewidth = 0.5) +
    scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1)) +
    labs(x = "position (Mbp)", y = "scaled (0-1)",
         title = sprintf("%s  (%d a-windows)", region_lab[[ch]], nrow(dw))) +
    theme_classic(base_size = 9) + theme(plot.title = element_text(size = 9))
}
```

```{r tracks-50, fig.height = 6.5}
(track_panel(ld_decay, "Chr1") / track_panel(ld_decay, "Chr4") /
   track_panel(ld_decay, "Chr6")) + plot_annotation(tag_levels = "a")
```

### How the decay track depends on the window count

The number of windows only controls the **smoothness** of the `a` track, not what
it says. With finer windows each `a` estimate rests on fewer marker pairs, so it
jitters more; min--max scaling then exaggerates whichever single window is most
extreme. Re-estimating with `n_win_decay = 100` makes this obvious -- the track is
spikier:

```{r decay-100, results = "hide"}
# keep_el = FALSE: we only need the windowed decay, not the edge lists
ld_decay_100 <- compute_LD_decay(gds, keep_el = FALSE, slide = 200,
                                 ld_method = "corr", n_win_decay = 100, cores = 2)
```

```{r tracks-100, fig.height = 6.5}
(track_panel(ld_decay_100, "Chr1") / track_panel(ld_decay_100, "Chr4") /
   track_panel(ld_decay_100, "Chr6")) + plot_annotation(tag_levels = "a")
```

Crucially, the spikes are **noise in the low-LD background only** -- they never
land inside an `ld_w` block. We can check this directly by pairing each window's
`a` with the mean `ld_w` of the markers it spans:

```{r robustness}
win_stats <- rbindlist(lapply(c("Chr1", "Chr4", "Chr6"), function(ch) {
  d <- as.data.table(ld_decay_100$by_chr[[ch]]$decay)[regime == "structured" & is.finite(a)]
  m <- map[Chr == ch]
  d[, win_ldw := vapply(seq_len(.N),
        function(i) mean(m[Pos >= start[i] & Pos <= end[i], ld_w_095], na.rm = TRUE),
        numeric(1))]
  d[is.finite(win_ldw), .(a, win_ldw)]
}))

# fast decay <-> low local LD
cor(win_stats$a, win_stats$win_ldw, method = "spearman")

# the a spikes (top-decile a) vs high-ld_w windows (top-decile ld_w):
top_a  <- win_stats$a      > quantile(win_stats$a,      0.90)
hi_ldw <- win_stats$win_ldw > quantile(win_stats$win_ldw, 0.90)
mean(hi_ldw[top_a])            # fraction of a-spikes falling in high-ld_w windows
```

The window-level `a`--`ld_w` correlation is strongly negative and **none** of the
`a` spikes fall in high-`ld_w` windows. So the two measures -- one at window
resolution, one at marker resolution -- independently agree on where the high-LD
regions are, and that agreement is invariant to the window count. This is why
`ld_w` is the statistic used to flag clusters for reconsideration during
complexity reduction: it carries the same signal at marker resolution, without
the windowing noise.

## Two-stage LD-complexity reduction

Stage 1 (`ld_complexity_reduction()`) clusters markers within each chromosome from
a sliding-window LD graph. In extended-LD regions like the Chr1 inversion, pairs
farther apart than the window are never compared, so a single block can fragment
into many adjacent clusters. Stage 2 (`ld_prune_and_eMLG()`) reconsiders the
flagged clusters directly from genotypes and consolidates them.

```{r stage1, results = "hide"}
stage1 <- ld_complexity_reduction(map = map, LD_decay = ld_decay, rho = 0.5, cores = 2)

result <- ld_prune_and_eMLG(
  GTs = GTs, stage1 = stage1, ld_w_col = "ld_w_095", ld_w_threshold = 0.05,
  LD_decay = ld_decay, rho = 0.95, score_threshold = 0.80, min_r2 = 0.2,
  min_n_loci_flag = 5, min_n_loci_eMLG = 5, compute_unflagged_eMLG = FALSE, cores = 2
)
```

`plot_pruning_comparison()` stacks Stage 1 against the combined result for one
chromosome. On the Chr1 inversion the effect is stark: the fragmented Stage-1
clusters (top) are consolidated into a few coherent groups (bottom).

```{r stage-plot, fig.height = 6.5}
# plot_pruning_comparison() returns its plot invisibly (and saves to out_folder),
# so assign and re-print it to show it here.
p_stage <- plot_pruning_comparison(chr = "Chr1", pruned_stage1 = stage1,
                                   result = result, map = map, out_folder = tempdir())
p_stage
```

### The Stage-2 merge distance window

Stage 2 only consolidates clusters whose consecutive physical gap stays below
`distance_threshold`. Left at its default (`NULL`), that threshold is **not** a
fixed constant: it is derived *per chromosome* from the decay model as
`d_from_rho(a_pred, rho)` -- the physical distance at which LD has decayed to the
relative threshold `rho` (default 0.95) -- using the chromosome-size-corrected
predicted decay rate `a_pred` rather than each chromosome's own noisy fit. On this
data it works out to only a few kb per region, and it adapts to each region's
decay:

```{r dist-default}
ds <- copy(ld_decay$decay_sum)
ds[, distance_threshold_bp := round(d_from_rho(a_pred, 0.95))]
ds[, .(Chr, a_pred = signif(a_pred, 3), distance_threshold_bp)]
```

A common alternative is a single fixed window -- e.g. 500 kb, as in the
Formica-hybrid analysis, whose low-recombination blocks span megabases:

```{r stage-500, results = "hide"}
result_500kb <- ld_prune_and_eMLG(
  GTs = GTs, stage1 = stage1, ld_w_col = "ld_w_095", ld_w_threshold = 0.05,
  LD_decay = ld_decay, rho = 0.95, score_threshold = 0.80, min_r2 = 0.2,
  min_n_loci_flag = 5, min_n_loci_eMLG = 5, compute_unflagged_eMLG = FALSE,
  distance_threshold = 5e5, cores = 2
)
```

```{r stage-500-plot, fig.height = 6.5}
p_500 <- plot_pruning_comparison(chr = "Chr1", pruned_stage1 = stage1,
                                 result = result_500kb, map = map, out_folder = tempdir())
p_500
```

On this compact example the two windows give almost the same partition:

```{r dist-compare}
c(default_rho_derived = nrow(result$groups),
  fixed_500kb         = nrow(result_500kb$groups))
```

The reason is that the flagged clusters inside a dense LD block are already
physically contiguous, so the cap rarely binds -- widening it from a few kb to
500 kb bridges few additional gaps. (In the 500 kb plot the inversion looks like
one green mass, but with hundreds of groups the colour palette simply recycles;
the group *count* confirms the partition barely changed.) The distance window
matters much more on data with *large* low-recombination blocks -- megabase
inversions, centromeric regions -- where a tight per-chromosome window can leave a
single block fragmented that a 500 kb window would consolidate.

Ideally the window would be defined in **genetic** rather than physical distance:
supplying a `genetic_map` and a single `cM_threshold` makes the merge window a
fixed number of centimorgans. Because a centimorgan spans *more* base pairs where
recombination is suppressed, a fixed cM window automatically widens (in bp)
precisely in the low-recombination regions that harbour the large LD blocks, and
narrows in the recombining bulk -- the behaviour you want, and one no single fixed
bp value can reproduce. Stickleback has no recombination map here, so we fall back
to physical distance.

Among the physical-distance options, raising `rho` widens the window. The default
`rho = 0.95` gave a few kb above; `rho = 0.99` is 5.2x wider
(`(0.99/0.01) / (0.95/0.05)`):

```{r dist-099}
ds[, `:=`(rho_0.95_bp = round(d_from_rho(a_pred, 0.95)),
          rho_0.99_bp = round(d_from_rho(a_pred, 0.99)))]
ds[, .(Chr, rho_0.95_bp, rho_0.99_bp)]
```

Even at `rho = 0.99` that reaches only ~40-85 kb here, still well short of 500 kb,
because these stickleback regions decay fairly fast (`a_pred ~1-3e-3`). Getting a
~500 kb window out of `rho = 0.99` would take about ten-fold slower decay
(`a_pred ~2e-4`) -- exactly the regime of large low-recombination blocks (e.g. the
megabase ant blocks the Formica analysis capped at 500 kb). So what "500 kb" means
is entirely relative to the local decay rate, which is why a cM-based window is the
principled choice where a map exists, with the `rho`-derived per-chromosome bp
window as the next-best adaptive proxy and a single fixed bp value the most
convenient but least adaptive option.

## Consensus fidelity and the flagging threshold

A Stage-1 cluster is flagged for Stage-2 merging when its local LD support exceeds
`ld_w_threshold`. Raising the threshold flags fewer clusters. The consensus
fidelity `score_eMLG = cor(round(x), x)^2` measures whether a merged block's
consensus still sits close to a discrete genotype; `ld_prune_and_eMLG()` enforces
a floor of 0.80 when accepting a merge. Sweeping the flagging threshold shows the
trade-off:

```{r sweep, results = "hide"}
TH <- c(0.025, 0.05, 0.1, 0.15)
fl <- rbindlist(lapply(TH, function(th) {
  r <- ld_prune_and_eMLG(GTs = GTs, stage1 = stage1, ld_w_col = "ld_w_095",
                         ld_w_threshold = th, score_threshold = 0.80, min_r2 = 0.2,
                         LD_decay = ld_decay, rho = 0.95, min_n_loci_flag = 5,
                         min_n_loci_eMLG = 5, compute_unflagged_eMLG = FALSE, cores = 2)
  g <- as.data.table(r$groups)[startsWith(group_id, "F")]
  g[, .(th = th, score, n_loci)]
}))
```

```{r fidelity-plot, fig.height = 6.5}
fl[, th_label := sprintf("ld_w threshold = %s  (n = %s, mean = %.3f)",
                         th, format(.N, big.mark = ","), mean(score)), by = th]
fl[, th_label := factor(th_label, levels = unique(th_label[order(th)]))]
fl[, cluster_type := fifelse(n_loci == 1, "singleton (no merge possible)", "merged (n_loci > 1)")]
hb <- fl[, { br <- seq(0.80, 1, by = 0.01); h <- hist(score, breaks = br, plot = FALSE)
             .(bin_mid = h$mids, count = h$counts) }, by = .(th_label, cluster_type)]
hb <- fl[, .(N = .N), by = th_label][hb, on = "th_label"][, pct := 100 * count / N]

ggplot(hb, aes(bin_mid, pct, fill = cluster_type)) +
  geom_col(colour = "white", linewidth = 0.1, width = 0.01) +
  geom_vline(xintercept = 0.80, linetype = 2, colour = "#D08A45", linewidth = 0.5) +
  scale_fill_manual(values = c("singleton (no merge possible)" = "#C7CDD2",
                               "merged (n_loci > 1)" = "#315B7D"), name = NULL) +
  facet_wrap(~ th_label, ncol = 1) + coord_cartesian(xlim = c(0.80, 1)) +
  labs(x = expression(score[eMLG] == cor(round(x), x)^2),
       y = "% of that run's flagged clusters") +
  theme_classic(base_size = 9) +
  theme(legend.position = "top", strip.text = element_text(size = 7.5, face = "bold"))
```

Raising `ld_w_threshold` shrinks the flagged set but the excluded clusters are the
fragmented extended-LD ones that would benefit from merging; lowering it pulls in
isolated single markers with no merge partner (the mass at the upper fidelity
limit). The threshold governs *which clusters are evaluated*, not which merges are
accepted -- the fidelity and correlation gates do that -- so a permissive setting
costs computation rather than accuracy.

## Association mapping: which cluster representation?

The complexity-reduction output feeds directly into association mapping. The
pruned representative markers give a genomic relationship matrix (GRM), and each
LD cluster can then be tested against a phenotype using one of three unit
representations: its **representative** SNP, its **eMLG consensus**, or its **best
single-SNP proxy** (`eMLG_best_snp()`, the member most correlated with the
consensus). Here we test ecotype (Marine vs Freshwater) with `emmax()` -- a
mixed-model association test that corrects for relatedness via the GRM (Li et al.
2018) -- and compare the F statistic across the three representations for the eMLG
clusters.

```{r assoc-setup, results = "hide"}
pheno <- stickleback$pheno
eco   <- as.integer(pheno$ecotype == "Marine")          # binary phenotype

# GRM from the pruned representative markers. autosome.only = FALSE because the
# Chr codes here are character labels, not integer autosome numbers.
K <- SNPRelate::snpgdsGRM(gds, snp.id = result$pruned, method = "GCTA",
                          autosome.only = FALSE, verbose = FALSE)$grm

best <- eMLG_best_snp(result, GTs, round_fill = FALSE)  # round_fill = FALSE keeps dosages

g <- as.data.table(result$groups)[has_eMLG == TRUE]
g[, best_marker := best$stats$best_marker[match(group_id, best$stats$group_id)]]

# EMMAX F per cluster, three representations, one shared GRM
Fr <- emmax(eco, GTs[, g$representative], K)$F    # representative SNP
Fc <- emmax(eco, result$eMLG[, g$group_id], K)$F  # eMLG consensus
Fm <- emmax(eco, GTs[, g$best_marker],     K)$F   # best single-SNP proxy
```

```{r assoc-plot, fig.height = 4}
fdt  <- data.table(Chr = g$Chr, representative = Fr, consensus = Fc, max_SNP = Fm)
long <- melt(fdt, id.vars = c("Chr", "consensus"), variable.name = "single_SNP",
             value.name = "F_single")
ggplot(long, aes(consensus, F_single, colour = Chr)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
  geom_point(alpha = 0.7, size = 1.2) +
  facet_wrap(~ single_SNP) +
  labs(x = "EMMAX F : eMLG consensus", y = "EMMAX F : single SNP",
       title = "Ecotype association F by cluster representation") +
  theme_bw(base_size = 10)
```

```{r assoc-summary}
# among the strongest-associated clusters, which representation gives the highest F?
top <- fdt[order(-consensus)][1:20]
table(apply(top[, .(representative, consensus, max_SNP)], 1,
            function(x) c("representative", "consensus", "max_SNP")[which.max(x)]))

# max F per chromosome
fdt[, lapply(.SD, function(x) round(max(x), 1)),
    by = Chr, .SDcols = c("representative", "consensus", "max_SNP")]
```

At the strongly-associated blocks the **consensus** usually gives the *highest* F
-- the opposite of the sorting statistic, where the consensus diluted a sharp
per-population fixation signal (see the `module_di25` analyses). The reason is the
nature of the statistic: ecotype association is a coherent, block-wide
*quantitative* signal, so averaging the cluster's markers into the consensus
**improves the signal-to-noise ratio**, whereas a sharp per-population fixation
call is carried by individual markers that averaging blurs. Here the effect is
clearest on the Chr1 inversion (a broad, coherently-associated block); at the more
localized Chr4 *Eda* locus the three representations are comparable. So the best
unit depends on the downstream statistic -- `eMLG_best_snp()` recovers signal for
sharp per-SNP measures, while the consensus is preferable for diffuse, block-wide
association -- and having all three on hand from one clustering lets you choose per
analysis.

### Sharper scans: LOCO GRM and ld_w prioritization

Two refinements make the single-SNP versus cluster comparison fairer and more
powerful.

**Leave-one-chromosome-out (LOCO) GRM.** The GRM above includes the chromosome
being tested, so it partly corrects away the signal it should detect. A LOCO GRM
tests each chromosome against relatedness estimated from the *other* chromosomes'
representatives, removing that proximal over-correction.

**`ld_w` prioritization.** If selection leaves a local-LD footprint, high-`ld_w`
markers are enriched for real signal. Restricting the single-SNP scan to the top
10% of `ld_w` (~1,600 of ~16,000 markers) concentrates testing on candidate
regions *and* greatly relaxes the FDR correction.

```{r loco, results = "hide"}
# LOCO GRMs: each chromosome vs relatedness estimated from the OTHER chromosomes
pruned_chr <- map$Chr[match(result$pruned, map$marker)]
chrs  <- unique(map$Chr)
Kloco <- setNames(lapply(chrs, function(ch)
  SNPRelate::snpgdsGRM(gds, snp.id = result$pruned[pruned_chr != ch],
                       method = "GCTA", autosome.only = FALSE, verbose = FALSE)$grm), chrs)

# (1) single-SNP LOCO scan, all markers
p_snp <- setNames(rep(NA_real_, nrow(map)), map$marker)
for (ch in chrs) { mk <- map[Chr == ch, marker]; p_snp[mk] <- emmax(eco, GTs[, mk], Kloco[[ch]])$pval }
map[, q_all := p.adjust(p_snp, "fdr")]

# (2) ld_w-prioritized: top 10% of ld_w only (FDR over that reduced set)
map[, top_ldw := ld_w_095 >= quantile(ld_w_095, 0.90, na.rm = TRUE)]
map[top_ldw == TRUE, q_top := p.adjust(p_snp[marker], "fdr")]

# (3) cluster-consensus LOCO scan
g[, p_clust := NA_real_]
for (ch in chrs) { gi <- g$Chr == ch
  if (any(gi)) g$p_clust[gi] <- emmax(eco, result$eMLG[, g$group_id[gi]], Kloco[[ch]])$pval }
g[, q_clust := p.adjust(p_clust, "fdr")]
```

```{r loco-compare}
data.frame(
  strategy      = c("all SNPs", "top 10% ld_w", "consensus clusters"),
  n_tests       = c(nrow(map), sum(map$top_ldw), nrow(g)),
  n_significant = c(sum(map$q_all < 0.05, na.rm = TRUE),
                    sum(map$q_top < 0.05, na.rm = TRUE),
                    sum(g$q_clust < 0.05))
)
```

LOCO alone sharpens the single-SNP scan: the Chr4 *Eda* peak now clears
significance where the whole-genome GRM had masked it. The `ld_w`-prioritized scan
gains the most hits, because the top-10% `ld_w` set is dominated by the large,
high-LD Chr1 inversion -- a coherent ecotype-differentiated block -- so relaxing
the FDR over that enriched subset lets most of it through. That is also the honest
caveat: high `ld_w` in an inversion reflects suppressed recombination as much as
selection, so the prioritized set concentrates on *candidate* regions rather than
proving selection at each marker.

The Manhattan below is the LOCO single-SNP scan, with SNPs in FDR-significant
consensus clusters coloured by `plot_pruning_comparison()`'s palette and rule (so
a coloured cluster carries the colour it has there):

```{r manh, fig.height = 3.2, fig.width = 8}
pal <- c("#E41A1C","#377EB8","#4DAF4A","#984EA3","#FF7F00","#FFFF33",
         "#A65628","#F781BF","#1B9E77","#D95F02","#7570B3","#66A61E")
allg <- as.data.table(result$groups)[startsWith(group_id, "F")]
allg[, col := pal[(rowid(Chr) - 1) %% length(pal) + 1]]
sig  <- allg[group_id %in% g[q_clust < 0.05, group_id]]
snp2col <- rbindlist(lapply(seq_len(nrow(sig)),
                            function(i) data.table(marker = sig$members[[i]], col = sig$col[i])))
map[, sig_col := snp2col$col[match(marker, snp2col$marker)]]
map[, in_sig := !is.na(sig_col)]
map[, neglogq := -log10(q_all)]

ggplot() +
  geom_hline(yintercept = -log10(0.05), linetype = 2, colour = "grey60") +
  geom_point(data = map[in_sig == FALSE], aes(Pos / 1e6, neglogq),
             colour = "grey75", size = 0.5, alpha = 0.5) +
  geom_point(data = map[in_sig == TRUE], aes(Pos / 1e6, neglogq, colour = sig_col), size = 1.3) +
  scale_colour_identity() +
  facet_wrap(~ Chr, scales = "free_x", nrow = 1) +
  labs(x = "position (Mbp)", y = expression(-log[10](q) ~ ", LOCO single-SNP"),
       title = "LOCO single-SNP scan; SNPs in FDR-significant consensus clusters coloured") +
  theme_bw(base_size = 9)
```

The three strategies lay out the trade-off. Single-SNP scanning localizes to the
marker (the Chr4 *Eda* signal is a clean per-SNP peak) but pays the full
multiple-testing cost. `ld_w` prioritization keeps SNP resolution while spending
tests only where local LD -- the selection/structure footprint -- is high.
Cluster-based scanning pools each coherent block into one powerful, low-resolution
test (the Chr1 inversion clusters), recovering block-wide signal a per-SNP scan
would split across many corrected tests.

```{r cleanup, include = FALSE}
SNPRelate::snpgdsClose(gds)
unlink(gds_path)
```

## Summary

On real stickleback data this worked example showed the full workflow --
per-chromosome LD-decay estimation, marker-resolution local LD support (`ld_w`),
and two-stage complexity reduction into a pruned marker set and eMLG consensus --
and demonstrated that the window-resolution decay rate and marker-resolution
`ld_w` agree, robustly, on where the high-LD regions are. It then fed the pruned
set into `emmax()` association mapping and compared cluster representations: for a
diffuse, block-wide quantitative signal like ecotype association the eMLG
consensus is preferable (averaging boosts signal-to-noise), whereas for a sharp
per-SNP measure the best single-SNP proxy (`eMLG_best_snp()`) recovers signal the
consensus would blur -- and one clustering yields all three representations, so
the choice can be made per analysis.
