---
title: "Calling outlier regions: the current LDscnR workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calling outlier regions: the current LDscnR workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4, dpi = 100,
  cache = TRUE, autodep = TRUE
)
```

```{r packages}
library(LDscnR)
library(data.table)
library(SNPRelate)
```

## What problem this solves

A genome scan tests many markers, but nearby markers on the same chromosome are
often correlated with each other because they are inherited together --
they are in **linkage disequilibrium (LD)**. That is a statement about how
alleles co-occur in a sample; it is not the same thing as **physical
linkage** (simply being close together on a chromosome), and it is not the
same thing as being under selection. LD can be strong between markers a
few hundred base pairs apart, or, around a large inversion or in a region
with little recombination, between markers many megabases apart. When
several correlated markers all cross a significance threshold, that is not
several independent pieces of evidence -- it is one signal, counted several
times.

`LDscnR` addresses this in two separate steps that this vignette walks
through in order:

1. Group markers into **phenotype-blind clusters** based only on how
   correlated they are with each other (never on the phenotype or the
   association result). Test one thing per cluster instead of one thing per
   marker.
2. After testing, assemble the clusters that turned out significant into
   **reported regions** -- genomic intervals a reader can look up and
   compare to other studies.

A separate, later step asks a question the first two cannot answer on their
own: could a plausible amount of population structure, on its own, produce
about this many discoveries? That needs a phenotype null built to preserve
whatever structure is a real concern for the dataset at hand -- there is no
generic answer, and this vignette builds one for its own example rather
than treating the choice as a formality.

This vignette runs the whole workflow on `stickleback`, a bundled subset of
a real three-spined stickleback panel (117 fish, ~16,000 markers, 3
chromosomes) contrasting marine and freshwater ecotypes (`?stickleback`).
Chromosome 1 carries a well-documented marine-freshwater inversion, which
gives this example something real to recover.

```{r cores}
cores <- 2
```

## Stage 1: phenotype-blind clustering

```{r data}
data("stickleback")
GTs <- stickleback$GTs
map <- copy(stickleback$map)
pheno <- copy(stickleback$pheno)
```

The clustering step only ever sees genotypes and marker positions -- not
`pheno` -- which is what makes it possible to build these groups once and
reuse them for any phenotype, and why a discovery later on cannot be an
artefact of having chosen the clusters to fit the answer.

```{r gds}
gds_path <- tempfile(fileext = ".gds")
gds <- create_gds_from_geno(GTs, map, gds_path)
```

```{r ld-decay, results = "hide"}
set.seed(1)
ld_decay <- compute_LD_decay(gds, keep_el = TRUE, slide = 200, n_win_decay = 20, cores = cores)
```

`ld_complexity_reduction()` fits a per-chromosome LD-decay curve (above) and
uses it to group markers into clusters -- see
`vignette("LDscnR_quick_introduction")` for the full mechanics. Each
cluster becomes one **Stage-1 unit**: a set of markers that are correlated
with each other on this dataset, not a claim that they are the same
biological haplotype, and not a claim that they are statistically
independent of every other unit (neighbouring units can still be somewhat
correlated; the clustering reduces redundancy, it does not certify
independence).

```{r stage1, results = "hide"}
stage1 <- ld_complexity_reduction(map = map, LD_decay = ld_decay, rho = 0.5, cores = cores)
```

```{r stage1-show}
stage1
```

## A relationship matrix built the same, phenotype-blind way

Association testing needs a kinship/relationship matrix to account for
relatedness and population structure. Consistent with the clustering above,
it is built from one representative marker per Stage-1 unit (`stage1$pruned`)
rather than every marker -- redundant, strongly-correlated markers would
otherwise be over-weighted in the relationship estimate.

```{r grm}
K <- snpgdsGRM(gds, snp.id = stage1$pruned, method = "GCTA",
               autosome.only = FALSE, verbose = FALSE)$grm
rownames(K) <- colnames(K) <- rownames(GTs)
```

## One p-value per Stage-1 unit

A cluster is not itself something an association test can consume -- it has
to become one variable per individual first. `ld_unit_matrix()` builds that
variable several ways (see `?ld_unit_matrix`); this vignette shows the two
most common:

- **Consensus testing**: average the genotypes within a unit into one
  value per individual, and test that value directly -- one association
  test per unit.
- **Simes combination**: test every marker individually as usual, then
  combine the p-values within each unit into a single, size-adjusted
  p-value.

Both are legitimate; they trade off differently between diluting
SNP-specific signal (consensus) and being penalised for unit size (Simes).
This package does not run the association test itself -- any engine that
returns p-values works. Here we use `emmax_fast()`, LDscnR's own EMMAX
implementation, for both.

```{r size-floor}
size_floor <- 8L   # illustrative floor for this three-chromosome subset
```

Eight is not a general cutoff. The full three-spined analysis in the
manuscript set its floor of eight from the density of the complete filtered
marker panel (approximately one required member per 100,000 markers), before
examining associations or EcoPeak overlap, and then checked sensitivity to
the floor. This subset is used to demonstrate the workflow, not to reselect
that full-panel setting.

```{r phenotype}
eco <- as.integer(pheno$ecotype == "Marine")
```

### Consensus testing

```{r consensus-test}
m_cons <- ld_unit_matrix(GTs, stage1, map, size_floor = size_floor, repr = "consensus_dosage")
prep_cons <- emmax_setup(m_cons, K)
p_unit <- emmax_fast(prep_cons, eco)
names(p_unit) <- colnames(m_cons)   # names/order must match attr(m_cons, "units")$unit_id
```

### Simes combination

```{r simes-test}
prep_marker <- emmax_setup(GTs, K)
p_marker <- emmax_fast(prep_marker, eco)
names(p_marker) <- colnames(GTs)   # names/order must match map$marker
```

`p_unit` is one p-value per tested unit; `p_marker` is one p-value per
marker, which `ld_outlier_test()` combines into a per-unit Simes p-value
internally. Either can be handed to `ld_outlier_test()` unnamed, in the
documented order, or named as above -- a named vector in the wrong order is
rejected with an error rather than silently scored against the wrong unit
or marker.

## BH correction and Stage-2 assembly

`ld_outlier_test()` applies Benjamini-Hochberg correction **at the unit
scale** -- one test per Stage-1 unit, not one per marker -- and then
assembles whichever units survive into reported regions (Stage 2). This
assembly step can only make significant units easier to read as intervals;
it cannot change which units were significant, because BH has already run
by the time it happens.

```{r outlier-test-consensus}
out_unit <- ld_outlier_test(stage1, map, p_unit, statistic = "unit", size_floor = size_floor,
                            alpha = 0.05, assembly = "physical")
out_unit
```

At the conventional `alpha = 0.05`, nothing here clears Benjamini-Hochberg
correction: the strongest unit sits on the known chromosome-1 inversion
(21.68-21.71 Mb) but its BH-adjusted q is about 0.086. That is the honest
result for this particular subset -- real, independently known biology can
still miss a chosen significance threshold in a modest sample, and a
vignette that only ever showed a clean discovery would be misleading about
that. For the rest of this vignette, to show what a positive result and
the machinery around it actually look like, we relax `alpha` to 0.10 --
**purely for illustration, not a general recommendation** -- which recovers
exactly one significant unit, on the inversion:

```{r outlier-test-illustrative}
out_unit <- ld_outlier_test(stage1, map, p_unit, statistic = "unit", size_floor = size_floor,
                            alpha = 0.10, assembly = "stage2_discovered",
                            GTs = GTs, LD_decay = ld_decay)
out_unit
out_unit$regions
```

The Simes route, tested independently from marker-level p-values, agrees
that this is the strongest signal in the dataset (though it need not
produce an identical q at every unit -- it is a different combination
rule, not a recomputation of the same number):

```{r outlier-test-simes}
out_simes <- ld_outlier_test(stage1, map, p_marker, statistic = "simes", size_floor = size_floor,
                             alpha = 0.10, assembly = "stage2_discovered",
                             GTs = GTs, LD_decay = ld_decay)
out_simes$units[significant == TRUE]
```

`assembly = "stage2_discovered"` (used above) re-clusters only the
significant units using the same LD rule Stage 1 used, so a reported
region can only span genuinely discovered signal. `assembly = "physical"`
is a simpler, distance-only merge, useful mainly as a cheap check that the
two rules broadly agree:

```{r outlier-test-physical}
out_phys <- ld_outlier_test(stage1, map, p_unit, statistic = "unit", size_floor = size_floor,
                            alpha = 0.10, assembly = "physical")
out_phys$regions
```

## Is the observed discovery count unusual?

One significant region is a result, not yet an answer to "could this be
explained by population structure alone?" That needs a **null built for
this dataset's actual structure**, not a generic one. Here, the four
sampling localities differ in what fraction of each is marine vs.
freshwater (`table(pheno$pop_locality, pheno$ecotype)`), so a phenotype
null that shuffles ecotype **within locality** asks the relevant question:
if ecotype carried no information beyond locality membership, how many
discoveries would this same pipeline make?

```{r perm-design}
table(pheno$pop_locality, pheno$ecotype)
```

B surrogates are drawn once, up front, into a fixed matrix rather than
generated fresh inside each `ld_outlier_perm()` call. That matters here
because we compare the unit-level and region-level counts below: a
function re-invoked separately by each call would consume different
random draws for each, making that comparison two independent
experiments rather than the same B surrogates counted two ways.

```{r perm-run}
B <- 40L
set.seed(1)
p_perm <- vapply(seq_len(B), function(b) {
  eco_perm <- ave(eco, pheno$pop_locality, FUN = sample)
  emmax_fast(prep_cons, eco_perm)
}, numeric(ncol(m_cons)))

null_units <- ld_outlier_perm(out_unit, stage1, map, p_perm,
                              level = "units", verbose = FALSE)
null_units
```

`null_units$null_obs_ratio` is a **null-to-observed discovery-count
ratio** -- how many discoveries the within-locality-shuffled surrogates
make, on average, relative to the one this analysis found -- not a
false-discovery rate, and not a statement about whether this particular
region is real. A small ratio here says the observed count is unusual
relative to this null; it does not, by itself, certify any one reported
region, and a differently-specified null (a different grouping, a
continuous structure axis) could reasonably give a different answer.

Counting at the **region** level instead of the unit level need not agree
with the unit-level count, since assembly can merge several significant
units into one region or leave them separate. Reusing the same `p_perm`
matrix keeps this a genuinely paired comparison -- both counts come from
the same B surrogate phenotypes, not two independently-drawn sets:

```{r perm-regions}
null_regions <- ld_outlier_perm(out_unit, stage1, map, p_perm, GTs = GTs, LD_decay = ld_decay,
                                level = "regions", verbose = FALSE)
null_regions
identical(null_units$surrogates, null_regions$surrogates)
```

In this particular run the two surrogate counts turn out identical, draw
for draw: with a signal this sparse, no surrogate ever produces more than
one significant unit, so there is never a chance for assembly to actually
merge anything, and the region count trivially tracks the unit count.
That is a property of this small example, not a general guarantee --
with several simultaneously significant units, a surrogate could easily
produce a different region count than unit count (units merged into
fewer regions, or vice versa across chromosomes), which is exactly why
both levels are computed rather than assumed to agree.

## A separate question: does this overlap external evidence?

Everything above asks whether the **discovery count** is unusual. A
completely different question is whether the **reported intervals**
land on external evidence -- candidate regions from an independent study,
a gene list, anything -- more often than their width alone would predict.
That is what `ld_region_rotation()` answers, and it is not a substitute for
the null above or vice versa: a region can pass one check and not the
other.

The annotation below is illustrative, not real published data -- three
made-up candidate intervals, one of which deliberately overlaps the
inversion region found above and two of which do not, so the mechanics are
visible on a dataset this small.

```{r rotation}
chrom_lengths <- map[, .(len = max(Pos)), by = Chr]
annotation <- data.table(
  Chr   = c("Chr1",    "Chr1",    "Chr4"),
  start = c(21690000,  5000000,   8000000),
  end   = c(21700000,  5100000,   8100000)
)
rotation <- ld_region_rotation(out_unit$regions, annotation, chrom_lengths,
                               scheme = "within", n_rotations = 2000, seed = 1)
rotation
```

The rotation relocates each reported region to a uniformly random position
that still fits on its own chromosome, keeping its width exactly as
observed, many times over, and asks how often a randomly placed region of
that same width would have landed on the annotation by chance. It is a
random-relocation null (each region placed independently), not a claim
that the annotation-overlap fold measures the same thing as the
discovery-count null above.

## Summary

- Stage 1 (`ld_complexity_reduction()`) groups markers into phenotype-blind
  units.
- `ld_unit_matrix()` turns a unit into one testable variable (consensus
  dosage) or leaves markers as-is for Simes combination -- either way,
  `ld_outlier_test()` applies BH correction once, at the unit scale.
- Stage 2 assembles significant units into reported regions, after
  correction, purely for reporting.
- `ld_outlier_perm()` asks whether the observed discovery count is unusual
  under a null built to preserve this dataset's own structure -- a
  null-to-observed ratio and a count-based p-value, not a per-region
  statement and not a false-discovery rate.
- `ld_region_rotation()` asks a different question: do the reported
  intervals overlap external evidence more than their width alone would
  predict.

```{r cleanup, include = FALSE}
SNPRelate::snpgdsClose(gds)
unlink(gds_path)
```
