---
title: "LD-aware outlier regions: calling and benchmarking"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LD-aware outlier regions: calling and benchmarking}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 3.2, dpi = 100,
  cache = TRUE, autodep = TRUE       # cache chunks so re-knits skip the LD-decay / null steps
)
```

```{r packages}
library(LDscnR)
library(data.table)
library(ggplot2)
library(SNPRelate)

# house style: white panels, no grey facet-strip background
theme_set(theme_bw(base_size = 10) + theme(strip.background = element_blank()))
```

## The idea

Genome scans for selection produce noisy single-SNP signals. LDscnR turns them
into **outlier regions (ORs)** by exploiting a single fact: **selection locally
inflates linkage disequilibrium**. A sweep or a barrier to gene flow raises LD
around its target, so the true signal is spread across a *block* of correlated
markers -- which is just another way of saying that true outliers **cluster** in
the genome. "LD-supported" and "clustered" are not two separate properties but the
same footprint of local LD, seen per-SNP and per-region. The pipeline scores each
SNP by how **consistently** it is both LD-prioritised and significant, calibrates a
threshold against a **structure-aware null**, and clusters the survivors into
regions.

The design collapses the usual thicket of tuning parameters to a principled few:

| role | parameter(s) | handled by |
|------|--------------|-----------|
| **swept / integrated** (natural [0,1]) | `rho` (`ld_w` window), `q*` (local-LD quantile) | folded into the C-score `ld_cscore()` |
| **fixed, principled** | `alpha` = 0.05 | the structure-aware null prices in any miscalibration |
| **clustering, LD-informed** | `rho_ld` (r²-link; lower = stricter LD separation), 500 kb cap | `ld_edges()` |
| **operational knobs** | `tau_C` (null-calibrated -- the hard one), `l_min` (null-floored post-filter; raise to focus) | `calibrate_tauc()` / `calibrate_lmin()` |

The rest of the vignette shows both halves: **benchmarking** a scan against
simulated truth (`sim_ex`), and **calling** ORs on real data with a
null-calibrated threshold (`stickleback`).

## Part 1 -- benchmarking on simulated data

`sim_ex` is a small three-chromosome simulation with two QTNs and precomputed
EMMAX and LFMM F-statistics (`?sim_ex`). We turn the F-statistics into p-values,
flag the detectable true-positive QTNs, and build the LD ingredients.

```{r sim-setup, results = "hide"}
data(sim_ex, envir = environment())
GTs <- sim_ex$GTs
map <- copy(sim_ex$map)
colnames(GTs) <- map$marker   # align GTs colnames ("1:Pos") to map markers ("Chr1:Pos"), same order
n   <- nrow(GTs)

# F -> p (df 1, n-2); MAF and per-chromosome Va share for the truth flag
map[, emx_p  := pf(emx_F,  1, n - 2, lower.tail = FALSE)]
map[, lfmm_p := pf(lfmm_F, 1, n - 2, lower.tail = FALSE)]
af <- colMeans(GTs) / 2
map[, MAF := pmin(af, 1 - af)[marker]]
map[, Va  := p_Va]
map <- flag_true_qtns(map, va_col = "Va", maf_col = "MAF", maf_min = 0.1, p_va_min = 0.05)

# per-marker max r^2 to any QTN (used by the diagnostic Manhattans below)
R2_qtn <- suppressWarnings(cor(GTs, GTs[, map[type == "QTN", marker], drop = FALSE],
                               use = "pairwise.complete.obs")^2)
R2_qtn[!is.finite(R2_qtn)] <- 0
map[, max_LD_to_QTN := apply(R2_qtn, 1, max)[marker]]

# LD decay + a small ld_w window grid (the `rho` axis of the C-score)
gds <- create_gds_from_geno(GTs, map, tempfile(fileext = ".gds"))
set.seed(1)
ld_decay <- compute_LD_decay(gds, keep_el = TRUE, slide = 50, ld_method = "corr",
                             n_win_decay = 20, cores = 2)
rho_grid <- c(0.5, 0.7, 0.85, 0.95)
ld_ws <- sapply(rho_grid, function(r) compute_ld_w(ld_decay, rho = r, cores = 2))
rownames(ld_ws) <- unlist(lapply(ld_decay$by_chr, function(o) o$snp_ids), use.names = FALSE)
colnames(ld_ws) <- as.character(rho_grid)
ld_ws <- ld_ws[map$marker, ]
decay_sum <- ld_decay$decay_sum
```

### The C-score, clustering, and TP/FP

`ld_cscore()` integrates over the `ld_w` window (`rho`) and the local-LD quantile
(`q*`) at fixed `alpha`: a SNP's score is the fraction of cells where it is both a
local-LD candidate and FDR-significant. High-C SNPs are then clustered into ORs
(`ld_edges()` builds the r²-graph once; `ld_regions()` clusters any gated subset),
and scored against the QTN truth with `evaluate_ors()` (dedup-neutral: a duplicate
region for an already-claimed QTN is counted as neither TP nor FP).

```{r sim-score}
C_emx <- ld_cscore(map$emx_p, ld_ws)                       # per-SNP consistency
edges <- ld_edges(names(C_emx)[C_emx > 0], GTs, map, decay_sum, rho_ld = 0.9)
th    <- score_thresholds(decay_sum, rho_r2 = 0.75, rho_d = 0.95, dmax_cap = 5e5)
qtab  <- qtn_ld_table(GTs, map, names(C_emx)[C_emx > 0], max_bp = 2e6)

regions <- ld_regions(names(C_emx)[C_emx >= 0.3], edges)     # gate at tau_C = 0.3
regions <- regions[lengths(regions) >= 2]                    # l_min = 2
ev <- evaluate_ors(regions, map, qtab, th$r2min, th$dmax)
unlist(ev[c("TP", "FP", "Recall", "Precision")])
```

The package ships a plotting helper, `ld_manhattan()`, that colours a per-marker
value (here the C-score) either by outlier region / group or by a continuous column.
Below we colour each region by whether it matches a true QTN (from `classify_ors()`)
and mark the QTNs with crosses: a region on a cross is a **true positive**, one away
from any cross is a **false positive**.

```{r sim-manhattan, fig.height = 2.6}
cl  <- classify_ors(regions, map, qtab, th$r2min, th$dmax)
grp <- unlist(lapply(seq_along(regions), function(i)
  setNames(rep(if (cl$is_TP[i]) "TP" else "FP", length(regions[[i]])), regions[[i]])))

ld_manhattan(map, value = setNames(as.numeric(C_emx), names(C_emx)), value_label = "C-score",
             group = grp, group_colours = c(TP = "#1F78B4", FP = "#E31A1C"),
             qtn = map[type == "QTN", marker], hline = 0.3,
             title = "Simulation: TP (blue) vs FP (red) regions; QTNs = crosses")
```

The two QTNs sit ~470 kb apart on Chr3 and fall inside one LD-linked region, so the
scan returns a single true-positive region (recall 0.5 under the dedup-neutral
metric -- one region can claim only one QTN). This is the *merge* direction of the
same distance/LD trade-off explored on the real data in Part 2; tightening the
distance cap here does not recover the second QTN (its independent signal is weak),
it only over-fragments the first QTN's block -- e.g. a 100 kb cap yields three
region fragments for the same two QTNs.

#### Why the C-score works: signal *and* local LD

`ld_manhattan()`'s continuous mode makes the logic visible. On the left, the
association statistic (`F`) coloured by local LD (`ld_w`): the real peaks on Chr3 are
both high-`F` **and** high-`ld_w`. On the right, `ld_w` coloured by each marker's LD
to the nearest QTN: high-`ld_w` markers near the QTNs are the genuine block, but note
the high-`ld_w` markers on Chr3's right arm that carry **no** association (dark) --
local LD alone is not enough. The C-score fires only where *both* hold, which is why
it beats either signal alone.

```{r sim-diagnostics, fig.height = 2.6}
ldw <- setNames(ld_ws[, "0.95"], rownames(ld_ws))
ld_manhattan(map, value = "emx_F", value_label = "EMMAX F",
             colour = ldw, colour_label = "ld_w",
             qtn = map[type == "QTN", marker],
             title = "Association (F) coloured by local LD (ld_w)")
```

```{r sim-diagnostics2, fig.height = 2.6}
ld_manhattan(map, value = ldw, value_label = "ld_w (rho 0.95)",
             colour = "max_LD_to_QTN", colour_label = "max r2 to QTN",
             qtn = map[type == "QTN", marker],
             title = "Local LD (ld_w) coloured by LD to the nearest QTN")
```

### C-score vs a single-SNP alpha scan

Because the C-score is one *ordered* knob (`tau_C`), a standard precision--recall
curve applies. We compare it against the single-SNP baseline (sweep `alpha` on the
raw p-values, same clustering and scoring), and summarise each with `pr_auc()`.

```{r sim-prauc}
pr_curve <- function(score_fun, grid) {
  rbindlist(lapply(grid, function(t) {
    reg <- ld_regions(score_fun(t), edges); reg <- reg[lengths(reg) >= 2]
    e <- evaluate_ors(reg, map, qtab, th$r2min, th$dmax)
    data.table(recall = e$Recall, precision = if (is.na(e$Precision)) NA_real_ else e$Precision)
  }))
}
cscore  <- pr_curve(function(t) names(C_emx)[C_emx >= t], seq(0.05, 1, 0.05))[, method := "C-score"]
singleS <- pr_curve(function(a) map$marker[p.adjust(map$emx_p, "BH") < a],
                    c(0.001, 0.005, 0.01, 0.05, 0.1))[, method := "single-SNP"]
c(C_score = pr_auc(cscore$recall, cscore$precision),
  single_SNP = pr_auc(singleS$recall, singleS$precision))
```

The curves themselves (this tiny example has only two QTNs, so they are coarse and
noisy -- on a full simulation with many replicates they separate cleanly):

```{r sim-prauc-plot, fig.height = 2.8}
prc <- rbind(cscore, singleS)[is.finite(precision) & is.finite(recall)]
ggplot(prc, aes(recall, precision, colour = method)) +
  geom_path() + geom_point(size = 1.8) +
  scale_colour_manual(values = c("C-score" = "#E31A1C", "single-SNP" = "steelblue"), name = NULL) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "Recall", y = "Precision", title = "Precision-recall: C-score vs single-SNP")
```

### Calibrating a second method against EMMAX (genomic control)

LFMM has no cheap structure-aware null. `gc_map_tauc()` transfers a threshold
calibrated on EMMAX onto LFMM's C-scale by quantile-matching -- genomic control on
the C-score, so LFMM is operated at EMMAX's FDR anchor.

```{r sim-gc}
C_lfmm  <- ld_cscore(map$lfmm_p, ld_ws)
tau_lfmm <- gc_map_tauc(tau_C = 0.3, C_ref = C_emx, C_target = C_lfmm)
signif(c(EMMAX_tau_C = 0.3, LFMM_tau_C = tau_lfmm), 3)
```

### The structure-aware null, checked against ground truth

Part 2 calls `ld_outlier_regions()` on real data, where `tau_C` comes from a
**structure-aware null** ([structured_null()]) rather than from truth. We can verify
that machinery *here*, where the truth is known. `sim_ex` ships only association
statistics, so we synthesise a phenotype with a known architecture -- the two QTNs'
genetic value, a genetic **structure** axis (something for the null to correct
away), and noise -- and run the whole pipeline on it.

```{r sim-null}
qtn  <- map[type == "QTN", marker]
Ksim <- snpgdsGRM(gds, snp.id = map$marker[ld_ws[, "0.95"] < 0.05],   # low-ld_w (neutral) GRM
                  method = "GCTA", autosome.only = FALSE, verbose = FALSE)$grm
set.seed(7)
y_sim <- as.numeric(1.5 * scale(rowSums(GTs[, qtn, drop = FALSE])) +       # QTN genetic value
                    2.0 * scale(eigen(Ksim, symmetric = TRUE)$vectors[, 1]) + # structure axis
                    rnorm(n))                                             # noise

res_sim <- ld_outlier_regions(
  y = y_sim, GTs = GTs, K = Ksim, ld_ws = ld_ws, map = map, decay_sum = decay_sum,
  basis = "genetic", rho_ld = 0.9, l_min = 2, fdr = 0.05, B = 100, seed = 1)
res_sim
```

The null does its job truth-free: structure-only surrogates almost never make
regions (`n_null` ~ 0), so the region-level FDR collapses as `tau_C` rises and the
threshold lands where the observed regions clear it.

```{r sim-null-fdr}
res_sim$fdr_curve[res_sim$fdr_curve$n_obs > 0, ]
```

Knowing the truth, we can confirm the calibrated call is right -- the recovered
region is a genuine QTN region (precision 1). As before, the two QTNs sit ~470 kb
apart and merge into one region, so recall is 0.5.

```{r sim-null-eval}
qtab_null <- qtn_ld_table(GTs, map, names(res_sim$C)[res_sim$C > 0], max_bp = 2e6)
unlist(evaluate_ors(res_sim$regions, map, qtab_null, th$r2min, th$dmax)[c("TP", "FP", "Recall", "Precision")])
```

## Part 2 -- calling ORs on real data with a null-calibrated threshold

`stickleback` is a three-spined-stickleback subset with a marine/freshwater
ecotype phenotype and three regions (`?stickleback`): the **Chr1 inversion** and
the **Chr4 *Eda*** locus (positive controls), and **Chr6** background (negative
control). Here there is no ground truth, so instead of benchmarking we *call* ORs
with `ld_outlier_regions()`, which calibrates `tau_C` from the structure-aware
null.

The kinship should capture structure without absorbing the signal, so we build it
from **low-`ld_w` (LD-independent) markers** -- an all-marker GRM over-corrects and
erases localised signals like *Eda*.

The two operating knobs play different roles. **`tau_C` is the hard one to intuit**
-- what counts as "consistent enough" on the C-scale -- and that is exactly what the
**structured null is for**: `calibrate_tauc()` sets it truth-free (here it lands at
the grid floor 0.05 because the structure surrogates are so clean; on the simulation
null-demo above it was 0.65). **`l_min` is the softer, post-filtering knob** -- how
large a region must be to be worth reporting. `calibrate_lmin()` gives a principled
*floor* for it from the same null (the largest region structure-only noise ever
builds, +1): here the null never clusters more than **2** SNPs, so the floor is
**`l_min = 3`** (`l_min = "auto"`). And that floor has a nice external check -- the
number of regions on the negative-control **Chr6 drops to exactly zero at `l_min =
3`** (17 at l_min 1, 1 at l_min 2, 0 at l_min 3).

Above that floor, `l_min` is a **focusing** control, not an FDR device -- and the
floor itself is only a *lower bound*, because `calibrate_lmin()` is **anti-conservative
by construction**. The null assumes **every SNP is neutral**: it has no true positives,
so it cannot reproduce the *satellite* clusters that real selected loci throw off --
small extra correlated blocks in the neighbourhood of, and often **driven by**, the
true target (LD shadows, secondary sweeps, structure around the locus). The null
simply never makes those, so the size it certifies is smaller than the size real data
needs. Empirically the satellites concentrate on exactly the chromosomes carrying the
large true regions (here Chr1 and Chr4 have them; the null-quiet Chr6 has none) -- a
pattern no structure null can correct for. The recommendation is therefore to **raise
`l_min` above the floor to focus on the largest regions** -- the big driver loci --
rather than the smaller blocks they induce nearby. If a map looks noisy, *especially
when the big-region chromosomes also carry many small ones*, that is the signature to
raise it. We show both views below.

**Separate regions by LD, not distance.** The other clustering knob is the r²-link
threshold, set by `rho_ld`. A distance cap alone is what every scan already does; the
point of an *LD-aware* method is to let linkage decide where one region ends and the
next begins. At the permissive canonical `rho_ld = 0.9` (r²-link ≈ 0.14 here) the
Chr4 candidates all link into one sprawling region -- overmerging that makes the scan
look noisier than it should. Lowering `rho_ld` to **0.75** raises the link threshold
to r² ≈ 0.25, which breaks Chr4 into the several regions one expects there, **while
the Chr1 inversion -- whose internal LD is genuinely high -- stays a single region
even at that stricter threshold** (it holds together down to r² ≈ 0.5). That is LD,
not the physical cap, doing the separation:

| `rho_ld` | r²-link | Chr1 | Chr4 |
|---------:|--------:|-----:|-----:|
| 0.90 | 0.14 | 1 | 1 (overmerged) |
| **0.75** | **0.25** | **1 (inversion)** | **several** |
| 0.50 | 0.47 | 1 | fewer (thinned) |

We therefore keep the default 500 kb cap and set `rho_ld = 0.75` (the same value used
for the simulations). The hard distance cap is still there as a backstop, and `rho_d`
remains available to derive the cap from LD decay instead -- but here LD does the work.

```{r stickle-setup, results = "hide"}
data(stickleback, envir = environment())
sGTs <- stickleback$GTs
smap <- copy(stickleback$map)
eco  <- as.integer(stickleback$pheno$ecotype == "Marine")

sgds <- create_gds_from_geno(sGTs, smap, tempfile(fileext = ".gds"))
set.seed(1)
sdecay <- compute_LD_decay(sgds, keep_el = TRUE, slide = 200, ld_method = "corr",
                           n_win_decay = 50, cores = 2)
srho <- c(0.5, 0.7, 0.85, 0.95)
sld_ws <- sapply(srho, function(r) compute_ld_w(sdecay, rho = r, cores = 2))
rownames(sld_ws) <- unlist(lapply(sdecay$by_chr, function(o) o$snp_ids), use.names = FALSE)
colnames(sld_ws) <- as.character(srho); sld_ws <- sld_ws[smap$marker, ]

# neutral-background GRM: GCTA on the low-ld_w (LD-independent) markers
neutral <- smap$marker[sld_ws[, "0.95"] < 0.05]
K <- snpgdsGRM(sgds, snp.id = neutral, method = "GCTA", autosome.only = FALSE, verbose = FALSE)$grm
```

```{r stickle-call}
res <- ld_outlier_regions(
  y = eco, GTs = sGTs, K = K, ld_ws = sld_ws, map = smap, decay_sum = sdecay$decay_sum,
  basis = "genetic",              # confounding is genetic (ecotype ~ structure)
  rho_ld = 0.75, dcap = 5e5, l_min = "auto", fdr = 0.05, B = 50, seed = 1,
  keep_cache = TRUE               # keep null + edges so we can inspect calibrate_lmin
)
res

# the l_min the null chose, and the null's max-region-size distribution behind it
c(l_min_auto = res$params$l_min,
  null_max_region = max(vapply(res$null$C_surr, function(Csp)
    max(c(0, lengths(ld_regions(names(Csp)[which(Csp >= 0.05)], res$edges)))), numeric(1))))

# the same engine's per-SNP q-values (BH-FDR of the low-ld_w-GRM EMMAX p), for the
# -log10(q) Manhattan below
q_emx <- p.adjust(emmax_fast(emmax_setup(sGTs, K), eco), "BH")
```

`ld_outlier_regions()` ran the whole pipeline -- fast EMMAX, C-score, a genetic
structure-aware null, the null-driven `l_min`, `tau_C` calibrated to FDR ≤ 0.05, and
clustering -- and returned the surviving ORs with their chromosome, extent and peak
C. The positive controls dominate: the Chr1 inversion comes through as one big region
and *Eda* on Chr4 as several LD-separated regions, plus a handful of smaller blocks
that clear the null; Chr6 stays quiet.

We show the result two ways with `ld_manhattan()`, each surviving region in its own
colour from `default_cluster_colours()` (grey = not in a region). First the C-score
itself, with the calibrated `tau_C`; then the same regions on a `-log10(q)` Manhattan
(BH-FDR of the underlying EMMAX scan) -- **identical region colours**, so you can read
across from the C-score the pipeline thresholds to the familiar q-value view.

```{r stickle-plot, fig.height = 2.8}
ld_manhattan(smap, value = setNames(as.numeric(res$C), names(res$C)), value_label = "C-score",
             regions = res$regions, hline = res$tau_C,
             title = sprintf("%d outlier regions at tau_C = %.2f (l_min = %d, rho_ld = 0.75)",
                             length(res$regions), res$tau_C, res$params$l_min))
```

At the null floor `l_min = 3` the Chr1 inversion and the three Chr4 *Eda* regions
come through, but Chr1 and Chr4 also carry a scatter of small satellite clusters --
the selection-driven neighbourhood blocks discussed above (note Chr6 stays empty).
If that reads as too busy, **raise `l_min` to focus on the largest regions**. This is
a pure post-filter on the cached clustering (no recomputation), so it is instant:

```{r stickle-focus, fig.height = 2.8}
reg10 <- ld_regions(names(res$C)[res$C >= res$tau_C], res$edges)   # reuse cached edges
reg10 <- reg10[lengths(reg10) >= 10]
ld_manhattan(smap, value = setNames(as.numeric(res$C), names(res$C)), value_label = "C-score",
             regions = reg10, hline = res$tau_C,
             title = sprintf("Focused view: l_min = 10 -> %d regions (satellites dropped)",
                             length(reg10)))
```

Same scan, same `tau_C` -- only the size filter changed. The satellites are gone and
the four large regions (Chr1 inversion, three Chr4 blocks) remain.

The distance cap `dcap` is a second, complementary way to clean up strays. Look again
at the focused view: the largest Chr4 region still *sprawls* ~770 kb, because a handful
of SNPs that are in LD with the block but physically distant get chained in by
sub-500 kb steps. Shortening the cap breaks that chain -- below ~100 kb those strays
detach, become sub-`l_min` satellites and drop out, so the region **tightens (769 ->
47 kb) with no change in region count** (200 kb only does this halfway on these data).
Crucially the **structure null is not rebuilt**: `dcap` enters only the clustering, so
this is a fresh `ld_edges()` (a `cor()`, ~0.4 s) over the *cached* C-scores -- no new
EMMAX, no new surrogates.

```{r stickle-dcap, fig.height = 2.8}
edges100 <- ld_edges(res$null$universe, sGTs, smap, sdecay$decay_sum,   # re-cluster only:
                     rho_ld = 0.75, dcap = 1e5)                          # 500 kb -> 100 kb cap
reg100 <- ld_regions(names(res$C)[res$C >= res$tau_C], edges100)
reg100 <- reg100[lengths(reg100) >= 10]
ld_manhattan(smap, value = setNames(as.numeric(res$C), names(res$C)), value_label = "C-score",
             regions = reg100, hline = res$tau_C,
             title = "dcap = 100 kb (l_min = 10): strays detached, Chr4 region tightened")
```

The Chr1 inversion is unaffected -- its dense core has no gaps that wide, so it holds
as one region down to a 25 kb cap.

**Two roles, not three interchangeable cleaners.** It is tempting to treat all three
clustering knobs as ways to "remove strays", but they split into two jobs.
`rho_ld` and `dcap` are **boundary** knobs: they decide *where a region ends* (by LD
and by distance) and so set each region's **shape and extent**. Tightening them sheds
distant or weakly-linked strays off a block -- but a shed piece does not vanish, it
becomes its own small region. `l_min` is the only **removal** knob: it decides *what
is big enough to report*, and drops a shed piece only when the piece is smaller than
`l_min`. They compose cleanly, but with two provisos: `l_min` must exceed the size of
the buds you want gone (here they run to ~7 SNPs, so a clean map wants `l_min` ≈ 10,
not the floor of 3 -- at `l_min = 3`, tightening `dcap`/`rho_ld` just re-labels the
strays as extra small regions rather than removing them); and pushing the boundary
knobs too far eventually fragments the *real* blocks as well. In short: **`rho_ld` and
`dcap` set the shape of the regions you keep; `l_min` sets how many.**

Finally, the null-floor result on the familiar `-log10(q)` scale, with **identical
region colours** so you can read across from the C-score to the q-value view:

```{r stickle-plot-q, fig.height = 2.8}
neglogq <- setNames(-log10(pmax(q_emx[smap$marker], 1e-30)), smap$marker)
ld_manhattan(smap, value = neglogq, value_label = expression(-log[10](italic(q))),
             regions = res$regions, hline = -log10(0.05),
             title = "Same outlier regions on the -log10(q) Manhattan")
```

## Summary

- **`ld_cscore()`** integrates the two [0,1] nuisance parameters (`rho`, `q*`) at
  fixed `alpha` into one per-SNP consistency score.
- **`ld_edges()` / `ld_regions()`** cluster the survivors with an LD-decay-informed
  r² link and a 500 kb cap; **`l_min`** is a region-size post-filter.
- **`structured_null()` / `calibrate_tauc()`** set the hard-to-intuit `tau_C`
  truth-free; **`calibrate_lmin()`** gives a data-driven floor for the `l_min`
  post-filter (raise it to focus on the largest regions); **`gc_map_tauc()`**
  transfers the threshold to a second method.
- **`ld_manhattan()`** draws the faceted region / continuous-coloured Manhattans.
- **`ld_outlier_regions()`** chains all of the above into one call for real data.
- The **`evaluate_ors()` / `pr_auc()`** benchmark scores any caller against
  simulated truth on a common OR frame -- a method-agnostic gold standard.

See `?ld_outlier_regions` and the individual function help pages for the full
argument set.
