---
title: "A quick introduction to LDscnR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A quick introduction to LDscnR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

```{r packages}
library(LDscnR)
library(data.table)
```

## Overview

`LDscnR` estimates chromosome-specific linkage disequilibrium (LD) decay from genotype data, and uses that decay model to reduce a marker set to LD-independent representatives. The same clustering step can be used either to produce a pruned marker set (for a kinship/relatedness matrix, EMMAX's `K`, BayPass's `OMEGA`), or, from that same clustering, one consensus genotype per LD block (an "eMLG" -- expected multi-locus genotype -- for block-level analyses such as long-range LD or Ohta's D statistics).

This vignette walks through the full workflow on the bundled simulated dataset `sim_ex` (3 chromosomes, 900 markers, 125 individuals -- small enough that every step below runs in well under a second):

1.  estimating LD decay,
2.  converting between relative LD thresholds, physical distance, and expected $r^2$,
3.  per-marker local LD support (`ld_w`),
4.  reducing markers to LD-independent representatives (Stage 1),
5.  consolidating into pruned markers and/or eMLGs (Stage 2), and
6.  a diagnostic plot comparing the two stages.

```{r cores}
# Kept small and explicit throughout this vignette -- sim_ex is tiny, so
# parallelism buys nothing here. Scale this up for real data.
cores <- 2
```

## Example data

```{r data}
data("sim_ex")

map <- sim_ex$map
GTs <- sim_ex$GTs

# sim_ex's GTs columns and map$marker refer to the same markers in the same
# order, just under different naming conventions -- align them once, up front.
colnames(GTs) <- map$marker
```

`LDscnR` reads genotypes from a GDS file (via `SNPRelate`), so the first step is to create one from the genotype matrix:

```{r gds}
gds_path <- tempfile(fileext = ".gds")
gds <- create_gds_from_geno(geno = GTs, map, gds_path)
```

## Estimating LD decay

`compute_LD_decay()` fits a per-chromosome decay curve $r^2(d) = b + (c-b)/(1+ad)$, where $a$ controls the rate of decay, $b$ is background LD, $c$ is short-range LD, and $d$ is physical distance. `keep_el = TRUE` retains the pairwise LD edge list each chromosome was fit from, which later steps (`compute_ld_w()`, `ld_complexity_reduction()`) reuse rather than recomputing.

`n_win_decay` and `slide` are set below for this toy dataset's very short simulated chromosomes; for real data, the defaults are usually appropriate, or can be tuned via a cheap preliminary run on a marker subset (`max_SNPs_decay`) -- see `?compute_LD_decay`.

```{r ld-decay, results = "hide"}
ld_decay <- compute_LD_decay(
  gds,
  n_win_decay = 5,
  keep_el = TRUE,
  slide = 200,
  cores = cores
)
```

```{r ld-decay-show}
ld_decay
```

```{r ld-decay-plot}
plot(ld_decay, type = "summary")
plot(ld_decay, type = "recommendation", rho = 0.99)
plot(ld_decay, type = "chr", chr = "Chr2")
```

## Relative LD thresholds, distance, and expected LD

Two helpers convert between a relative threshold $\rho$, physical distance, and expected $r^2$:

- `d_from_rho(a, rho)` -- physical distance (bp) at which $\rho$ is reached, for a chromosome with decay rate `a`.
- `ld_from_rho(b, c, rho)` -- expected $r^2$ at $\rho$.

```{r rho-helpers}
chr1 <- ld_decay$decay_sum[1]

d_from_rho(a = chr1$a, rho = 0.99)
ld_from_rho(b = chr1$b, c = chr1$c, rho = 0.99)
```

These conversions underpin `ld_decay$recommendation`'s suggested window sizes, and the per-chromosome clustering thresholds `ld_complexity_reduction()` derives below.

## Per-marker local LD support

`compute_ld_w()` summarises, for each marker, how much LD support it has from nearby markers within the physical window implied by $\rho$. It accepts a vector of $\rho$ values and computes all of them in one pass per chromosome -- each chromosome's edge list is read/symmetrised once and reused, not re-read once per threshold.

```{r ld-w}
ld_w <- compute_ld_w(ld_decay, rho = c(0.90, 0.95, 0.99), cores = cores)
map[, ld_w_095 := ld_w[, "rho_0.95"]]
```

## Stage 1: reduce markers to LD-independent representatives

`ld_complexity_reduction()` clusters markers within each chromosome (connected components, then complete-linkage refinement within each) and picks one representative marker per cluster, at a chromosome-specific threshold derived from the decay model at the target pruning $\rho$. This single call is the shared starting point for both a pruning-only workflow and an eMLG workflow -- see below.

```{r stage1, results = "hide"}
stage1 <- ld_complexity_reduction(map = map, LD_decay = ld_decay, rho = 0.5, cores = cores)
```

```{r stage1-show}
stage1
```

If only a pruned marker set is needed (a GRM, EMMAX's `K`, BayPass's `OMEGA`), `stage1$pruned` is already a ready-to-use character vector of representative markers -- nothing further is required.

## Stage 2 (optional): consolidate and summarise as eMLGs

Stage 1's edge list is built from a sliding window, so a genuinely correlated, low-recombination block can still come out fragmented into several adjacent clusters (pairs farther apart than the window were never evaluated at all). `ld_prune_and_eMLG()` fixes this after the fact, but only for the clusters that need it: those flagged by high local LD support (`ld_w_col`/`ld_w_threshold`) are re-compared directly from genotypes -- with no window restriction -- and consolidated via a distance-restricted, quality-gated dynamic cut. This produces a refined pruned marker set and an eMLG matrix (one consensus genotype per block, useful where a single representative SNP would throw away most of a correlated block's information -- e.g. long-range LD or Ohta's D statistics) from the same pass. Unflagged clusters (usually the large majority) pass straight through unchanged.

`ld_w_threshold = 0` is used below purely because `sim_ex` has so few markers per chromosome that a realistic threshold (e.g. `0.2`, appropriate on real data) flags almost nothing here -- there simply isn't enough local marker density for `ld_w` to build up. `ld_w_threshold = 0` is not really "off": on real data, virtually every marker has *some* positive local LD support, so it routes essentially everything through the expensive merge step rather than skipping it -- see the speed tip below for the tuning that actually matters at realistic scale.

`distance_threshold` (the maximum physical gap allowed between clusters still considered part of one contiguous, mergeable block) is left at its default here, which derives a per-chromosome value from `rho` and `LD_decay` (`d_from_rho(a_pred, rho)`) rather than one fixed bp constant -- see `?ld_prune_and_eMLG` for why this uses each chromosome's size-corrected predicted decay rate rather than its own observed decay rate.

```{r stage2, results = "hide"}
result <- ld_prune_and_eMLG(
  GTs = GTs, stage1 = stage1, ld_w_col = "ld_w_095", ld_w_threshold = 0,
  LD_decay = ld_decay, rho = 0.95,
  score_threshold = 0.80, min_r2 = 0.2, cores = cores
)
```

```{r stage2-show}
# one row per final group (flagged clusters consolidated, unflagged passed through)
nrow(result$groups)

# refined pruned marker set
length(result$pruned)

# individuals x blocks consensus-genotype matrix
dim(result$eMLG)
```

### Speed tip: preliminary vs. final runs

`ld_prune_and_eMLG()`'s cost is dominated by an all-pairs correlation among the *flagged* clusters, which scales roughly quadratically with how many clusters get flagged. On real data, `ld_w_threshold` is the lever that matters, and it should move in different directions depending on the run:

- **Preliminary/exploratory runs**: use a high `ld_w_threshold` (or the dataset's default) to flag only the most obviously redundant clusters -- fast, good enough for a first look at cluster counts and eMLG behaviour.
- **Final run**: lower `ld_w_threshold` toward `~0.05` (optionally combined with `min_n_loci_flag`, to also pull in large-but-low-`ld_w` clusters) so that as many genuine mergers as possible actually happen -- slower, but this only needs to be run once.
- **`compute_unflagged_eMLG = FALSE`** skips eMLG computation for the unflagged clusters entirely (usually the large majority) if only the pruned marker set is needed, independent of `ld_w_threshold`.

## Comparing Stage 1 and Stage 2

`plot_pruning_comparison()` stacks Stage 1's raw clusters (top panel) against Stage 2's consolidated groups (bottom panel) for one chromosome, so fragmented blocks and their reunited counterparts can be compared directly -- both panels plot the same underlying markers, just grouped (and thus coloured) differently.

`ld_w_col`/`ld_w_threshold`/`min_n_loci_flag` default to `result$params`, i.e. whatever `ld_prune_and_eMLG()` was actually called with above, so the two panels can't silently end up comparing different flagging criteria (passing a value that disagrees with `result$params` warns -- the Combined panel only ever reflects `result`'s own flagging, so an inconsistent threshold typically makes it look empty rather than erroring). `out_folder` is set to a temporary directory here purely for this vignette; omit it to use the default (`"./Figures/"`).

```{r comparison-plot}
# plot_pruning_comparison() returns its plot invisibly (it also saves the
# figure to out_folder), so it needs to be assigned and re-printed to show
# up here.
p_comparison <- plot_pruning_comparison(
  chr = "Chr3", pruned_stage1 = stage1, result = result, map = map,
  out_folder = tempdir()
)
p_comparison
```

## Summary

Starting from genotype and SNP metadata, this vignette:

- created a GDS object and estimated chromosome-wise LD decay,
- converted between relative LD thresholds, physical distance, and expected $r^2$,
- computed per-marker local LD support across several $\rho$ values in one pass,
- reduced markers to LD-independent representatives (Stage 1), and
- consolidated flagged clusters into a refined pruned marker set and an eMLG matrix (Stage 2), with a diagnostic plot comparing the two stages.

For empirical data, the same workflow applies; only the tuning parameters (`slide`, `rho`, `ld_w_threshold`, ...) need adjusting to the dataset's scale.

```{r cleanup, include = FALSE}
SNPRelate::snpgdsClose(gds)
unlink(gds_path)
```
